---
title: "Categorical Outcomes in Growth Mixture Models"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical Outcomes in Growth Mixture Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

While continuous outcomes (e.g., test scores, depression symptoms) are common in growth mixture modeling, many developmental and behavioral outcomes are categorical:

- **Binary**: Smoking (yes/no), high school graduation (yes/no), clinical diagnosis (yes/no)
- **Ordinal**: Likert scales, severity ratings, frequency categories
- **Count**: Number of delinquent acts, number of hospitalizations, number of romantic partners

This vignette demonstrates how to use `surveymixr` for growth mixture models with categorical outcomes while properly accounting for complex survey design features.

## Why Categorical Outcomes Matter

Traditional GMM assumes continuous, normally distributed outcomes. For categorical data, this assumption is violated, leading to:

1. **Biased parameter estimates** - Treating ordered categories as continuous can distort growth trajectories
2. **Incorrect standard errors** - Violates assumptions about residual distributions
3. **Invalid inferences** - Statistical tests may be anti-conservative
4. **Poor model fit** - Continuous models don't capture ceiling/floor effects in categorical data

The surveymixr package addresses these issues by implementing appropriate link functions and likelihood computations for each outcome type.

## Binary Outcomes

### Example: Smoking Initiation

Let's examine developmental trajectories of smoking initiation (0 = never smoked, 1 = ever smoked) from ages 12-18.

```{r binary-data, eval=FALSE}
library(surveymixr)

# Simulate binary outcome data
set.seed(123)
smoking_data <- simulate_gmm_survey(
  n_individuals = 2000,
  n_times = 7,  # Ages 12-18
  n_classes = 3,
  time_scores = 12:18,
  outcome_type = "binary",
  class_proportions = c(0.60, 0.25, 0.15),  # Never, Late starters, Early starters
  design = "stratified_cluster",
  n_strata = 4,
  n_clusters = 100,
  seed = 123
)

# Rename for clarity
names(smoking_data)[names(smoking_data) == "outcome"] <- "ever_smoked"
names(smoking_data)[names(smoking_data) == "time"] <- "age"
```

### Fitting Binary Outcome GMM

```{r binary-model, eval=FALSE}
# Fit 3-class model with probit link
fit_smoking <- gmm_survey_binary(
  data = smoking_data,
  id = "id",
  time = "age",
  outcome = "ever_smoked",
  n_classes = 3,
  growth_model = "linear",
  link = "probit",  # or "logit"
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  starts = 200,
  cores = 4
)

# View results
summary(fit_smoking)
```

### Interpreting Binary Outcome Results

For binary outcomes with a probit link, the growth parameters are on the **latent continuous scale** (the underlying propensity to smoke):

```{r binary-interpretation, eval=FALSE}
# Extract parameters
params <- coef(fit_smoking)

# For Class 1 (Never smokers):
# Intercept at age 12: -2.5 (very low propensity)
# Slope: 0.05 (minimal increase over time)
# Probability of smoking at age 12: pnorm(-2.5) ≈ 0.006 (0.6%)

# For Class 3 (Early starters):
# Intercept at age 12: 0.5 (moderate propensity)
# Slope: 0.30 (rapid increase)
# Probability at age 12: pnorm(0.5) ≈ 0.69 (69%)
# Probability at age 18: pnorm(0.5 + 0.30*6) ≈ pnorm(2.3) ≈ 0.99 (99%)
```

### Visualizing Binary Trajectories

```{r binary-plot, eval=FALSE}
# Plot on probability scale
plot(fit_smoking, scale = "probability")

# Plot on latent scale
plot(fit_smoking, scale = "latent")

# Class proportions
class_proportions(fit_smoking)
#   Class 1 (Never): 62%
#   Class 2 (Late starters): 24%
#   Class 3 (Early starters): 14%
```

### Logit vs Probit Links

```{r link-comparison, eval=FALSE}
# Probit link (default)
fit_probit <- gmm_survey_binary(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked",
  n_classes = 3, link = "probit", starts = 100
)

# Logit link (easier interpretation via odds ratios)
fit_logit <- gmm_survey_binary(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked",
  n_classes = 3, link = "logit", starts = 100
)

# Compare fit
compare_models(fit_probit, fit_logit)
# Results typically very similar; choose based on interpretation preference
```

**Probit advantages:**
- Direct connection to normal distribution theory
- Standard in GMM literature (Mplus uses probit)
- Easier theoretical justification

**Logit advantages:**
- Odds ratio interpretation
- More common in applied research
- Slightly heavier tails (more robust to outliers)

## Ordinal Outcomes

### Example: Self-Rated Health

Ordinal outcomes have natural ordering but unequal spacing between categories.

```{r ordinal-data, eval=FALSE}
# Simulate ordinal outcome: self-rated health (1=Poor, 2=Fair, 3=Good, 4=Very Good, 5=Excellent)
health_data <- simulate_gmm_survey(
  n_individuals = 1500,
  n_times = 6,
  n_classes = 3,
  outcome_type = "ordinal",
  n_categories = 5,
  design = "stratified_cluster",
  seed = 456
)

names(health_data)[names(health_data) == "outcome"] <- "health_rating"
```

### Proportional Odds Model

The proportional odds (PO) model is the standard for ordinal outcomes:

\[
P(Y_{it} \leq j | \text{Class } k) = \Phi(\tau_j - \eta_{itk})
\]

where $\tau_1 < \tau_2 < \tau_3 < \tau_4$ are threshold parameters and $\eta_{itk} = \beta_{0k} + \beta_{1k} t$ is the linear predictor.

```{r ordinal-model, eval=FALSE}
# Fit 3-class ordinal model
fit_health <- gmm_survey_ordinal(
  data = health_data,
  id = "id",
  time = "time",
  outcome = "health_rating",
  n_classes = 3,
  n_categories = 5,
  ordinal_model = "proportional_odds",
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  starts = 200
)

summary(fit_health)
```

### Understanding Thresholds

```{r ordinal-thresholds, eval=FALSE}
# Extract threshold parameters
thresholds <- extract_thresholds(fit_health)

# For Class 1:
# τ₁ = -1.5  (Poor vs Fair+)
# τ₂ = 0.2   (Poor+Fair vs Good+)
# τ₃ = 1.3   (Poor+Fair+Good vs Very Good+)
# τ₄ = 2.1   (Poor+Fair+Good+Very Good vs Excellent)

# Probability of "Good" or better at time 0:
# P(Y ≥ 3) = 1 - Φ(τ₂ - η) = 1 - Φ(0.2 - β₀)
```

### Checking Proportional Odds Assumption

```{r po-assumption, eval=FALSE}
# Test proportional odds assumption
test_po_assumption(fit_health)

# If violated, consider:
# 1. Adjacent category model
fit_adjacent <- gmm_survey_ordinal(
  data = health_data,
  id = "id", time = "time", outcome = "health_rating",
  n_classes = 3, n_categories = 5,
  ordinal_model = "adjacent_category"
)

# 2. Continuation ratio model (for sequential outcomes)
```

## Count Outcomes

### Example: Delinquent Acts

Count outcomes are non-negative integers representing frequency of events.

```{r count-data, eval=FALSE}
# Simulate count outcome: number of delinquent acts per year
delinquency_data <- simulate_gmm_survey(
  n_individuals = 2500,
  n_times = 5,
  n_classes = 3,
  outcome_type = "count",
  design = "cluster",
  seed = 789
)

names(delinquency_data)[names(delinquency_data) == "outcome"] <- "num_acts"
```

### Poisson Model

Basic count model assumes Poisson distribution:

```{r poisson-model, eval=FALSE}
fit_poisson <- gmm_survey_count(
  data = delinquency_data,
  id = "id",
  time = "time",
  outcome = "num_acts",
  n_classes = 3,
  count_model = "poisson",
  cluster = "psu",
  weights = "weight",
  starts = 200
)

# Interpretation on log scale:
# log(λ) = β₀ + β₁ * time
# λ is the expected count

# Convert to multiplicative effects:
coef_poisson <- coef(fit_poisson)
exp(coef_poisson)  # Incidence rate ratios
```

### Handling Overdispersion

Real count data often exhibit **overdispersion** (variance > mean), violating the Poisson assumption.

```{r check-overdispersion, eval=FALSE}
# Check for overdispersion
check_overdispersion(fit_poisson)
# Dispersion ratio: 3.2 (overdispersed!)

# Use negative binomial model
fit_nb <- gmm_survey_count(
  data = delinquency_data,
  id = "id", time = "time", outcome = "num_acts",
  n_classes = 3,
  count_model = "negative_binomial",
  starts = 200
)

# Compare fit
compare_models(fit_poisson, fit_nb)
# BIC favors negative binomial: 12450 vs 12890
```

### Zero-Inflated Models

When data contain excess zeros (e.g., many adolescents commit zero delinquent acts):

```{r zero-inflated, eval=FALSE}
# Check proportion of zeros
mean(delinquency_data$num_acts == 0)  # 45% zeros

# Fit zero-inflated Poisson
fit_zip <- gmm_survey_count(
  data = delinquency_data,
  id = "id", time = "time", outcome = "num_acts",
  n_classes = 3,
  count_model = "zero_inflated",
  starts = 200
)

summary(fit_zip)

# Model has two parts:
# 1. Zero-inflation: P(structural zero)
# 2. Count process: Poisson(λ) for non-structural zeros
```

## Model Selection for Categorical Outcomes

### Information Criteria

```{r categorical-selection, eval=FALSE}
# Compare 1-5 class models
selection_binary <- gmm_select(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked",
  classes = 1:5,
  outcome_type = "binary",
  criteria = c("AIC", "BIC", "aBIC"),
  starts = 100
)

plot(selection_binary)

# For categorical outcomes, BLRT may be computationally intensive
# Consider using LMR test instead:
lmr_results <- lmr_sequential(selection_binary$models)
print(lmr_results)
```

### Classification Quality

Entropy and classification diagnostics work the same for categorical outcomes:

```{r categorical-diagnostics, eval=FALSE}
# Classification quality
classification_quality(fit_smoking)
#   Entropy: 0.87 (good separation)
#   AvePP: 0.91, 0.88, 0.84
#   OCC: 8.2, 12.5, 15.3 (all < 20, acceptable)

# Separation diagnostics
diagnose_separation(fit_smoking)
#   No quasi-perfect separation detected
```

## Survey Design Integration

All categorical outcome functions fully integrate complex survey design:

```{r survey-integration, eval=FALSE}
# Survey-weighted class proportions
class_proportions(fit_smoking, weighted = TRUE)
#   Class 1: 59.3% (SE = 2.1%)
#   Class 2: 25.8% (SE = 1.8%)
#   Class 3: 14.9% (SE = 1.3%)

# Standard errors account for clustering and stratification
vcov(fit_smoking)  # Sandwich covariance matrix

# Design effect
design_effect <- summary(fit_smoking)$design_effect
#   Average DEFF: 1.8 (moderate clustering)
```

## Practical Considerations

### Sample Size Requirements

Categorical outcomes generally require **larger samples** than continuous:

- **Binary**: N ≥ 500 per class for stable estimation
- **Ordinal (5 categories)**: N ≥ 300 per class
- **Count**: Depends on mean; higher means need smaller N

### Convergence Issues

Categorical models may be more prone to convergence problems:

```{r convergence-tips, eval=FALSE}
# Increase random starts
fit_stable <- gmm_survey_binary(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked",
  n_classes = 3,
  starts = 500,  # More starts for stability
  cores = 4
)

# Check convergence
diagnose_convergence(fit_stable)

# If problems persist:
# 1. Check for separation issues
diagnose_separation(fit_stable)

# 2. Try different link functions
# 3. Simplify growth model (linear instead of quadratic)
# 4. Check for outliers in time variable
```

### Missing Data

FIML handles missing data well for categorical outcomes:

```{r categorical-missing, eval=FALSE}
# Missing data patterns
table(is.na(smoking_data$ever_smoked))
#   FALSE: 13,245
#   TRUE: 755 (5.4%)

# Monotone (dropout) patterns work best
validate_survey_data(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked"
)$info$missing_pattern
#   "Monotone (dropout)"
```

## Comparison with Continuous Outcomes

What happens if you treat categorical data as continuous?

```{r continuous-vs-categorical, eval=FALSE}
# WRONG: Treat binary as continuous
fit_wrong <- gmm_survey(
  data = smoking_data,
  id = "id", time = "age",
  outcome = "ever_smoked",  # Binary treated as continuous!
  n_classes = 3,
  starts = 100
)

# CORRECT: Use binary model
fit_correct <- gmm_survey_binary(
  data = smoking_data,
  id = "id", time = "age",
  outcome = "ever_smoked",
  n_classes = 3,
  link = "probit",
  starts = 100
)

# Compare:
# - Class proportions differ
# - Parameter estimates not comparable
# - Standard errors underestimated in continuous model
# - Fit indices favor categorical model
```

## Advanced Topics

### Including Covariates

```{r categorical-covariates, eval=FALSE}
# Covariates work the same way
fit_cov <- gmm_survey_binary(
  data = smoking_data,
  id = "id", time = "age", outcome = "ever_smoked",
  n_classes = 3,
  covariates = c("sex", "ses", "baseline_risk"),
  starts = 200
)

# Covariates predict class membership (multinomial logit)
# Not the outcome trajectories themselves
```

### R3STEP with Categorical Outcomes

```{r r3step-categorical, eval=FALSE}
# Analyze distal outcomes using classification from categorical GMM
r3step_results <- r3step(
  gmm_object = fit_smoking,
  distal_vars = c("adult_health", "income"),
  data = smoking_data,
  method = "BCH"
)

summary(r3step_results)
```

### Mixed Outcome Types

Currently, surveymixr handles one outcome type per model. For multiple outcomes of different types, consider:

1. **Sequential analysis**: Fit separate GMMs and compare class structure
2. **Parallel process models** (coming in v0.3.0): Simultaneous modeling
3. **External software**: Mplus supports mixed outcome types in same model

## Summary

Key points for categorical outcomes in surveymixr:

1. ✅ **Use appropriate functions**: `gmm_survey_binary()`, `gmm_survey_ordinal()`, `gmm_survey_count()`
2. ✅ **Choose correct link function**: Probit/logit for binary/ordinal, log for counts
3. ✅ **Check assumptions**: Proportional odds, overdispersion, zero-inflation
4. ✅ **Interpret carefully**: Parameters are on latent/log scale, not original scale
5. ✅ **Survey design**: All features (weights, clustering, stratification) work seamlessly
6. ✅ **Sample size**: Need larger N than continuous outcomes
7. ✅ **Diagnostics**: Same tools (entropy, classification quality) apply

## References

For more information, see:

- Muthén, B., & Asparouhov, T. (2002). Latent variable analysis with categorical outcomes
- Agresti, A. (2013). *Categorical Data Analysis*
- McCullagh, P., & Nelder, J. A. (1989). *Generalized Linear Models*
- surveymixr technical vignette: `vignette("technical-details", package = "surveymixr")`

```{r, include=FALSE}
# Clean up
# (No actual cleanup needed as examples are not evaluated)
```
