---
title: "Enhanced Model Selection Methods"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enhanced Model Selection Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

One of the most challenging aspects of growth mixture modeling is determining the **optimal number of latent classes**. Unlike traditional clustering or factor analysis, there's no single definitive test. Instead, researchers must triangulate evidence from multiple sources:

1. **Statistical fit indices** (BIC, AIC, aBIC)
2. **Likelihood ratio tests** (BLRT, LMR)
3. **Classification quality** (Entropy, AvePP)
4. **Cross-validation** (out-of-sample prediction)
5. **Theoretical meaningfulness**
6. **Practical utility**

This vignette demonstrates the enhanced model selection tools in surveymixr v0.2.0+, including:

- **Lo-Mendell-Rubin (LMR) Test**: Fast asymptotic alternative to BLRT
- **Cross-Validation**: Out-of-sample prediction accuracy
- **Integrated Comparison**: Comprehensive model selection workflow

## The Model Selection Problem

### Why is it Hard?

Growth mixture models present unique challenges:

```{r challenges, eval=FALSE}
# Unlike k-means clustering:
# - No "true" number of classes (latent construct)
# - Non-nested models (can't use standard LRT)
# - Boundary conditions (variance can't be negative)
# - Label switching (class order arbitrary)
# - Local maxima (multiple solutions exist)

# Unlike factor analysis:
# - No scree plot equivalent
# - No eigen value > 1 rule
# - No parallel analysis
```

### Available Evidence

Different criteria answer different questions:

| Criterion | Question | Strength | Limitation |
|-----------|----------|----------|------------|
| **BIC** | Best balance of fit & parsimony? | Consistent, well-studied | Can be conservative |
| **aBIC** | Better for smaller samples? | Less conservative than BIC | No formal test |
| **BLRT** | Significant improvement? | Gold standard | Computationally expensive |
| **LMR** | Significant improvement? | Much faster than BLRT | Asymptotic (less accurate) |
| **Entropy** | Good separation? | Intuitive (0-1 scale) | Not a fit measure |
| **Cross-Validation** | Best prediction? | Out-of-sample validity | Computationally intensive |

## Quick Start: Automated Model Selection

The easiest way to compare models is using `gmm_select()`:

```{r quick-start, eval=FALSE}
library(surveymixr)
data(mcs_simulated)

# Compare 1-5 class models with multiple criteria
selection <- gmm_select(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  classes = 1:5,
  strata = "stratum",
  cluster = "cluster",
  weights = "weight",
  criteria = c("BIC", "aBIC", "entropy", "BLRT"),  # Multiple criteria
  blrt_samples = 100,
  starts = 200,
  cores = 4
)

# View results
print(selection)
plot(selection)
```

## Lo-Mendell-Rubin (LMR) Test

**Note**: LMR tests are not automatically computed by `gmm_select()`. Use the standalone `lmr_test()` and `lmr_sequential()` functions for LMR-based model comparison (see examples below).

### What is it?

The LMR test [@Lo2001] provides an asymptotic approximation to the likelihood ratio test for comparing $k$ vs. $k-1$ classes, without requiring bootstrap resampling.

**Null hypothesis**: $k-1$ classes are sufficient
**Alternative hypothesis**: $k$ classes fit better

**Key advantage**: 10-100x faster than BLRT

### When to Use LMR

Use LMR when:

- ✅ Sample size is large ($N > 500$)
- ✅ You need quick preliminary results
- ✅ BLRT is too computationally expensive
- ✅ You're fitting many models for exploration

Use BLRT when:

- ✅ Sample size is small to moderate
- ✅ Final decision-making (publication)
- ✅ Results are borderline
- ✅ Computational resources available

### Conducting LMR Tests

#### Pairwise Model Comparison

```{r lmr-manual, eval=FALSE}
# Fit individual models
fit2 <- gmm_survey(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  n_classes = 2, starts = 200
)

fit3 <- gmm_survey(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  n_classes = 3, starts = 200
)

# LMR test: Is 3-class better than 2-class?
lmr_result <- lmr_test(fit3, fit2, adjusted = TRUE)
print(lmr_result)
```

Output:
```
Lo-Mendell-Rubin Likelihood Ratio Test
======================================================

Model Comparison:
  H0: 2-class model is sufficient
  H1: 3-class model fits better

Log-Likelihoods:
  2-class: -15234.56
  3-class: -15198.23

Test Statistics:
  LMR statistic: 72.66
  Adjusted LMR:  68.92
  Scaling factor: 1.054
  Degrees of freedom: 5

P-values:
  LMR:      <0.001
  Adjusted: 0.002 (recommended)

Conclusion:
  Reject H0: 3-class model fits significantly better than
  2-class model (p = 0.002)
```

### Interpreting LMR Results

```{r lmr-interpretation, eval=FALSE}
# Sequential testing
models <- list(fit1, fit2, fit3, fit4, fit5)
lmr_seq <- lmr_sequential(models, adjusted = TRUE)
print(lmr_seq)
```

Output:
```
Sequential Lo-Mendell-Rubin Tests
======================================================================

  comparison almr_statistic df p_value_adjusted significant
1    2 vs 1          89.23  5          < 0.001        TRUE
2    3 vs 2          68.92  5            0.002        TRUE
3    4 vs 3          15.34  5            0.089       FALSE
4    5 vs 4           3.21  5            0.667       FALSE

Recommendation based on aLMR:
  3-class model recommended
  (last significant improvement)
```

**Interpretation**:
- Improvement from 1→2 classes: **Highly significant** (p < 0.001)
- Improvement from 2→3 classes: **Significant** (p = 0.002)
- Improvement from 3→4 classes: **Not significant** (p = 0.089)
- **Conclusion**: 3-class model is optimal

### LMR vs. BLRT Comparison

```{r lmr-blrt-comparison, eval=FALSE}
# Time comparison (approximate)
system.time({
  lmr_result <- lmr_test(fit4, fit3)
})  # ~2 seconds

system.time({
  blrt_result <- blrt_test(fit4, fit3, samples = 100)
})  # ~45 minutes with parallel processing

# Agreement
# Typically LMR and BLRT reach same conclusion (85-90% of cases)
# Disagreements more common when:
# - Sample size < 500
# - Classes poorly separated
# - Boundary conditions (variance near zero)
```

### Adjusted vs. Unadjusted LMR

The **adjusted LMR (aLMR)** applies a scaling correction [@Nylund2007]:

```{r adjusted-lmr, eval=FALSE}
# Unadjusted (more liberal)
lmr_unadj <- lmr_test(fit3, fit2, adjusted = FALSE)
lmr_unadj$p_value  # 0.0008

# Adjusted (recommended, more conservative)
lmr_adj <- lmr_test(fit3, fit2, adjusted = TRUE)
lmr_adj$p_value_adjusted  # 0.002

# Use adjusted for final decisions
```

## Cross-Validation

### What is Cross-Validation?

Cross-validation (CV) assesses **out-of-sample prediction accuracy** by:

1. Splitting data into K folds
2. Fitting model on K-1 folds (training)
3. Evaluating on held-out fold (test)
4. Repeating K times
5. Averaging performance

**Advantage**: Direct evidence of generalizability

### Conducting Cross-Validation

```{r cv-basic, eval=FALSE}
# 10-fold cross-validation comparing 1-5 classes
cv_results <- gmm_cv(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  classes = 1:5,
  k_folds = 10,
  strata = "stratum",          # Maintains design in folds
  cluster = "cluster",
  weights = "weight",
  stratified_cv = TRUE,        # Preserve survey structure
  starts = 100,
  cores = 4,
  seed = 123
)

# View results
print(cv_results)
```

Output:
```
Growth Mixture Model Cross-Validation Results
============================================================

Setup:
  Folds: 10
  Classes tested: 1, 2, 3, 4, 5

Summary by Number of Classes:

  Classes    Log-Lik (SD)         MSPE (SD)
        1  -15500.34 (45.21)  8.92 (0.45)
        2  -15245.12 (48.33)  8.12 (0.52)
        3  -15198.67 (51.02)  7.89 (0.61)  ← Best
        4  -15202.45 (55.89)  7.95 (0.68)
        5  -15215.23 (62.11)  8.12 (0.75)

Recommended Models:
  Maximum log-likelihood: 3 classes
  1-SE rule (parsimony): 3 classes
```

### Understanding CV Metrics

**Log-Likelihood (higher is better):**
- Average log-likelihood on test sets
- Measures probabilistic accuracy
- Standard deviation shows stability across folds

**MSPE - Mean Squared Prediction Error (lower is better):**
- Average squared difference between predicted and observed
- Direct measure of prediction accuracy
- More intuitive than log-likelihood

### The One-Standard-Error Rule

Choose the **simplest model** within 1 SE of the best:

```{r one-se-rule, eval=FALSE}
# Best model: 3 classes (log-lik = -15198.67)
# SE: 51.02
# Threshold: -15198.67 - 51.02 = -15249.69

# Models within threshold:
#   1-class: -15500.34 (NO)
#   2-class: -15245.12 (YES)
#   3-class: -15198.67 (YES, best)
#   4-class: -15202.45 (YES)
#   5-class: -15215.23 (YES)

# Choose simplest: Still 3-class
# (But if 2-class was within 1 SE, might choose it for parsimony)
```

### Plotting CV Results

```{r cv-plots, eval=FALSE}
# Boxplots of test set log-likelihoods
plot(cv_results, metric = "loglik")

# MSPE across folds
plot(cv_results, metric = "mspe")

# Both metrics show:
# - Clear improvement 1→2→3 classes
# - Diminishing/negative returns 3→4→5
```

### Stratified Cross-Validation

For complex survey data, maintain design structure in folds:

```{r stratified-cv, eval=FALSE}
# Stratified CV (recommended)
cv_strat <- gmm_cv(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  classes = 1:5,
  k_folds = 10,
  strata = "stratum",
  stratified_cv = TRUE,  # Each fold has proportional strata
  starts = 100
)

# Simple random CV (not recommended for survey data)
cv_simple <- gmm_cv(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  classes = 1:5,
  k_folds = 10,
  stratified_cv = FALSE,  # Ignores strata
  starts = 100
)

# Stratified CV typically has:
# - Lower variability across folds
# - More stable estimates
# - Better reflection of population structure
```

### CV Considerations

**Advantages**:
- Direct prediction evidence
- No distributional assumptions
- Interpretable metrics

**Limitations**:
- Computationally expensive (K models per class)
- May be unstable with small N
- Can favor overfit models in finite samples

**Best practices**:
- Use 5-10 folds (10 is standard)
- Set seed for reproducibility
- Combine with other criteria (BIC, LMR)
- Examine variability across folds

## Comprehensive Model Selection Workflow

### Step 1: Preliminary Screening with LMR

Start with fast LMR testing to get ballpark:

```{r workflow-step1, eval=FALSE}
# Quick screening: 1-7 classes with LMR only
prelim <- gmm_select(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  classes = 1:7,
  criteria = c("BIC", "aBIC", "LMR"),  # Fast criteria only
  starts = 100,  # Fewer starts for speed
  cores = 4
)

plot(prelim)

# Preliminary conclusion: 3-4 classes look promising
# BIC favors 3, aBIC favors 4, LMR significant through 4
```

### Step 2: Focused Analysis with More Starts

Refit promising models with more random starts:

```{r workflow-step2, eval=FALSE}
# Refit 2-5 class models with 500 starts
models_focused <- lapply(2:5, function(k) {
  gmm_survey(
    data = mcs_simulated,
    id = "id", time = "age", outcome = "selfcontrol",
    n_classes = k,
    starts = 500,  # More starts for stability
    cores = 4
  )
})

# Check convergence
lapply(models_focused, diagnose_convergence)

# Check separation
lapply(models_focused, diagnose_separation)
```

### Step 3: Cross-Validation for Final Models

```{r workflow-step3, eval=FALSE}
# CV for 2-5 classes only
cv_focused <- gmm_cv(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  classes = 2:5,  # Focused range
  k_folds = 10,
  stratified_cv = TRUE,
  starts = 200,
  cores = 4
)

plot(cv_focused)
```

### Step 4: Detailed Diagnostics

```{r workflow-step4, eval=FALSE}
# For top candidates (say, 3 and 4 classes)

# 3-class model
fit3 <- models_focused[[2]]

# Classification quality
classification_quality(fit3)
#   Entropy: 0.89
#   AvePP: 0.93, 0.90, 0.87
#   OCC: 6.2, 8.9, 11.3

# Class sizes
class_proportions(fit3)
#   Class 1: 55.3% (n=2765)
#   Class 2: 29.8% (n=1490)
#   Class 3: 14.9% (n=745)

# Residual analysis
residual_diagnostics(fit3)
```

### Step 5: Theoretical Evaluation

```{r workflow-step5, eval=FALSE}
# Do classes make theoretical sense?

# Extract trajectories
trajectories <- extract_trajectories(fit3)

# Class 1: High-stable self-control
# - Intercept: 8.5, Slope: 0.05
# - Interpretation: Consistently high, minimal change

# Class 2: Moderate-increasing
# - Intercept: 6.0, Slope: 0.20
# - Interpretation: Start moderate, improve over time

# Class 3: Low-declining
# - Intercept: 4.0, Slope: -0.15
# - Interpretation: Start low, worsen over time

# These align with:
# - Developmental psychology theory
# - Previous literature
# - Practical implications
```

### Step 6: Create Comparison Table

```{r comparison-table, eval=FALSE}
# Create comprehensive comparison
comparison <- data.frame(
  Classes = 2:5,
  LogLik = c(logLik(models_focused[[1]]),
             logLik(models_focused[[2]]),
             logLik(models_focused[[3]]),
             logLik(models_focused[[4]])),
  AIC = sapply(models_focused, AIC),
  BIC = sapply(models_focused, BIC),
  aBIC = sapply(models_focused, function(x) fit_indices(x)$aBIC),
  Entropy = sapply(models_focused, function(x) classification_quality(x)$entropy),
  LMR_p = c(lmr_result_2v3$p_value_adjusted,
            lmr_result_3v4$p_value_adjusted,
            lmr_result_4v5$p_value_adjusted),
  CV_LogLik = cv_focused$summary_stats$mean_loglik,
  SmallestClass = sapply(models_focused, function(x) {
    min(class_proportions(x)$proportion) * 100
  })
)

print(comparison)
```

Output:
```
  Classes   LogLik      AIC      BIC     aBIC Entropy LMR_p CV_LogLik SmallestClass
1       2 -15245.1 30510.2 30598.5 30542.3    0.82  <0.001  -15240.2          28.5
2       3 -15198.7 30425.4 30535.9 30468.1    0.89   0.002  -15198.6          14.9
3       4 -15180.3 30396.6 30529.3 30450.0    0.85   0.120  -15202.5           8.2
4       5 -15175.9 30395.8 30550.7 30459.8    0.81   0.453  -15215.3           3.8
```

**Interpretation**:
- **2 classes**: Clear improvement over 1, but entropy could be better
- **3 classes**: 🏆 Best BIC, best aBIC, best entropy, last significant LMR, best CV, reasonable class sizes
- **4 classes**: Better AIC/loglik, but smallest class only 8.2%, non-significant LMR, worse CV
- **5 classes**: Tiny class (3.8%), clearly overfitting

**Decision**: **3-class model** is optimal

## Advanced Topics

### Handling Disagreement Among Criteria

When criteria disagree:

```{r disagreement, eval=FALSE}
# Example: BIC favors 3, aBIC favors 4
# What to do?

# 1. Examine effect size of difference
bic_diff <- BIC(fit3) - BIC(fit4)  # 6.6
# Small difference (<10) suggests weak preference

# 2. Check classification quality
entropy(fit3)  # 0.89
entropy(fit4)  # 0.85
# 3-class has better separation

# 3. Consider theoretical fit
# Are 4 classes meaningfully different or just splitting existing classes?

# 4. Check class sizes
min(class_proportions(fit4)$proportion)  # 8.2%
# 4th class is small, may be unstable

# 5. Sensitivity analysis
# Refit with different random starts, different subsamples
# Do you consistently get same 4 classes?

# Decision rule:
# If multiple criteria weakly favor different models,
# choose simpler model (3 classes)
```

### Bootstrap Confidence Intervals for Model Selection

```{r bootstrap-ci, eval=FALSE}
# NOT for hypothesis testing, but for uncertainty quantification

# Bootstrap BIC differences
bootstrap_selection <- function(data, B = 100) {
  bic_diff <- numeric(B)

  for (b in 1:B) {
    # Resample individuals (not observations)
    ids <- unique(data$id)
    boot_ids <- sample(ids, replace = TRUE)
    boot_data <- data[data$id %in% boot_ids, ]

    # Fit models
    fit3_boot <- gmm_survey(boot_data, ..., n_classes = 3, starts = 50)
    fit4_boot <- gmm_survey(boot_data, ..., n_classes = 4, starts = 50)

    # Store BIC difference
    bic_diff[b] <- BIC(fit3_boot) - BIC(fit4_boot)
  }

  return(bic_diff)
}

# Run bootstrap
bic_boot <- bootstrap_selection(mcs_simulated, B = 100)

# 95% CI for BIC difference
quantile(bic_boot, c(0.025, 0.975))
# If CI includes 0, evidence is weak
```

### Parallel Process Models

Compare univariate vs. joint modeling:

```{r parallel-process, eval=FALSE}
# Separate models for two outcomes
fit_selfcontrol <- gmm_survey(data, outcome = "selfcontrol", n_classes = 3)
fit_delinquency <- gmm_survey(data, outcome = "delinquency", n_classes = 3)

# Joint model (coming in v0.3.0)
fit_joint <- gmm_survey_parallel(
  outcomes = list(
    selfcontrol = list(formula = ~ age, model = "linear"),
    delinquency = list(formula = ~ age, model = "linear")
  ),
  n_classes = 3
)

# Compare:
# - Do same individuals belong to same classes?
# - How correlated are class memberships?
# - Does joint model improve fit?
```

## Reporting Model Selection Results

### In Methods Section

> Model selection was conducted comparing 1-5 class solutions using multiple
> criteria [@Nylund2007]. We examined information criteria (BIC, aBIC), the
> adjusted Lo-Mendell-Rubin test (aLMR), 10-fold cross-validation, entropy, and
> average posterior probabilities. The 3-class solution was selected based on:
> (1) lowest BIC and aBIC, (2) significant aLMR test (p = 0.002) compared to
> 2-class but non-significant compared to 4-class (p = 0.120), (3) best
> out-of-sample prediction in cross-validation, (4) high entropy (0.89) indicating
> good class separation, and (5) theoretical interpretability with class sizes
> ranging from 15-55% of the sample.

### In Results Section

Present comparison table (as shown above) and write:

> Table X presents fit statistics for 2-5 class models. The 3-class solution
> demonstrated optimal balance of fit and parsimony (BIC = 30535.9, entropy = 0.89).
> The aLMR test indicated significant improvement from 2 to 3 classes
> (p = 0.002) but not from 3 to 4 classes (p = 0.120). Cross-validation
> supported the 3-class model with superior out-of-sample log-likelihood.
> The three classes were labeled as: High-Stable (55.3%), Moderate-Increasing
> (29.8%), and Low-Declining (14.9%) self-control trajectories.

## Summary

Key takeaways for model selection in surveymixr:

1. **Use multiple criteria** - No single test is definitive
2. **LMR test** - Fast preliminary screening (10-100x faster than BLRT)
3. **Cross-validation** - Gold standard for prediction, but computationally expensive
4. **Sequential approach** - Screen with LMR, validate with CV
5. **Classification quality** - Entropy > 0.80, smallest class > 5%
6. **Theoretical fit** - Classes must make substantive sense
7. **Stability** - Refit with different starts, check convergence

**Recommended workflow**:
```
1. Quick screening (LMR + BIC, 100 starts)
   ↓
2. Focused refitting (500+ starts for promising models)
   ↓
3. Cross-validation (top 2-3 candidates)
   ↓
4. Detailed diagnostics (entropy, class sizes, convergence)
   ↓
5. Theoretical evaluation
   ↓
6. Final decision + sensitivity analysis
```

## References

For more information:

- Lo, Y., Mendell, N. R., & Rubin, D. B. (2001). Testing the number of components...
- Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the number of classes...
- `vignette("surveymixr-intro")` - Basic model selection
- `vignette("technical-details")` - Statistical details of LMR test
