---
title: "Analyzing Distal Outcomes with R3STEP"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Distal Outcomes with R3STEP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(surveymixr)
```

## Introduction

After identifying latent classes using growth mixture modeling, researchers often want to examine how classes differ on **distal outcomes** - variables measured after or separate from the trajectory analysis.

**Examples of distal outcomes**:
- Delinquency (after self-control trajectories)
- Academic achievement (after engagement trajectories)
- Health outcomes (after stress trajectories)
- Employment status (after skill development trajectories)

**The Problem**: Simply assigning individuals to their most likely class and comparing means is **biased** due to **classification error** [@bakk2013estimating].

**The Solution**: The R3STEP method accounts for classification uncertainty, providing unbiased estimates of class differences on distal outcomes.

## The Three Steps

### Step 1: Fit Unconditional Growth Mixture Model

First, fit a GMM **without** the distal outcome. This ensures class formation isn't influenced by the distal variable.

```{r step1, eval=FALSE}
# Load example data
data(mcs_simulated)

# Fit 3-class model of self-control trajectories
gmm_fit <- gmm_survey(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  n_classes = 3,
  growth_model = "linear",
  strata = "stratum",
  cluster = "cluster",
  weights = "weight",
  starts = 500,
  cores = 4
)

summary(gmm_fit)

# Create person-level data for R3STEP analysis
# R3STEP requires one row per individual
person_data <- mcs_simulated[!duplicated(mcs_simulated$id), ]
```

### Step 2: Assign Most Likely Class

Extract posterior probabilities and assign each individual to their most likely class:

```{r step2, eval=FALSE}
# Get posterior probabilities (stored in fitted model)
posterior_probs <- gmm_fit@posterior_probs

# Most likely class assignment
most_likely_class <- apply(posterior_probs, 1, which.max)

table(most_likely_class)
```

### Step 3: Analyze Distal Outcome

Use the R3STEP method to test for class differences on the distal outcome, accounting for classification error.

```{r step3, eval=FALSE}
# Analyze delinquency as distal outcome
r3_result <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "delinquency",
  data = person_data,
  method = "BCH"  # Recommended method
)

summary(r3_result)
```

## Methods Implemented

`surveymixr` implements three R3STEP methods:

### 1. BCH Method (Recommended)

The Bolck-Croon-Hagenaars [@bolck2004estimating] method uses weights to correct for classification error.

**Advantages**:
- Computationally efficient
- Works well with survey weights
- Provides survey-adjusted standard errors
- Most widely used

**When to use**: Default choice for most applications

```{r bch, eval=FALSE}
result_bch <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "academic_achievement",
  data = person_data,
  method = "BCH"
)

# View class-specific means
result_bch@class_means

# View omnibus test (do classes differ overall?)
result_bch@test_results$omnibus_test

# View pairwise comparisons
result_bch@test_results$pairwise_tests
```

### 2. ML Method

Maximum likelihood approach jointly models class membership and distal outcome.

**Advantages**:
- Theoretically optimal
- Handles complex relationships

**Disadvantages**:
- More computationally intensive
- May be unstable with small classes

**When to use**: When BCH assumptions are violated or for publication comparisons

```{r ml, eval=FALSE}
result_ml <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "delinquency",
  data = person_data,
  method = "ML",
  cores = 4  # Can parallelize
)

summary(result_ml)
```

### 3. Manual 3-Step

For custom analyses or when you want full control over Step 3.

```{r manual, eval=FALSE}
result_manual <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "academic_achievement",
  data = person_data,
  method = "manual"
)

# Returns data with class assignments and weights
# Researcher performs own analysis in Step 3
```

## Interpreting Results

### Class-Specific Means

```{r means, eval=FALSE}
# Extract class means
means_table <- result_bch@class_means

print(means_table)
#   class  mean    se  lower_ci upper_ci   n
#   1      2.1  0.15     1.8      2.4    550
#   2      4.3  0.18     3.9      4.7    300
#   3      7.2  0.22     6.8      7.6    150
```

**Interpretation**:
- Class 1 (high self-control): Lowest delinquency (mean = 2.1)
- Class 2 (moderate): Moderate delinquency (mean = 4.3)
- Class 3 (low self-control): Highest delinquency (mean = 7.2)

### Omnibus Test

Tests whether classes differ on the distal outcome (overall F-test or Wald test).

```{r omnibus, eval=FALSE}
omnibus <- result_bch@test_results$omnibus_test

print(omnibus)
# $statistic: 145.6
# $df1: 2
# $df2: 997
# $p_value: < 0.001
```

**Interpretation**:
- p < 0.001: Classes differ significantly on delinquency
- Reject null hypothesis of equal means

### Pairwise Comparisons

Which specific classes differ from each other?

```{r pairwise, eval=FALSE}
pairwise <- result_bch@test_results$pairwise_tests

print(pairwise)
#   comparison  diff    se  statistic  p_value  sig
#   1 vs 2      -2.2  0.24    -9.2     <.001    ***
#   1 vs 3      -5.1  0.27   -18.9     <.001    ***
#   2 vs 3      -2.9  0.29   -10.0     <.001    ***
```

**Interpretation**: All three classes differ significantly from each other on delinquency.

### Effect Sizes

```{r effect_sizes, eval=FALSE}
# Cohen's d for pairwise comparisons
effect_sizes <- result_bch@effect_sizes

print(effect_sizes)
#   comparison  cohens_d  magnitude
#   1 vs 2         0.65    medium
#   1 vs 3         1.42    large
#   2 vs 3         0.78    medium-large
```

## Visualizing Results

### Plot Class Means

```{r plot_means, eval=FALSE}
plot(result_bch, type = "means")
```

This produces a bar plot with:
- Class-specific means
- 95% confidence intervals
- Significance indicators

### Plot Distributions

```{r plot_distributions, eval=FALSE}
plot(result_bch, type = "distributions")
```

Shows density plots or violin plots of the distal outcome by latent class.

## Practical Examples

### Example 1: Continuous Distal Outcome

**Research Question**: Do self-control trajectory classes differ in academic achievement?

```{r example1, eval=FALSE}
# Step 1: Fit GMM (already done above)

# Step 2+3: Analyze achievement as distal
achievement_result <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "academic_achievement",
  data = person_data,
  method = "BCH"
)

summary(achievement_result)

# Plot results
plot(achievement_result)

# Export table for publication
library(gt)
gt_table <- gt(achievement_result@class_means) %>%
  tab_header(title = "Academic Achievement by Self-Control Trajectory Class") %>%
  fmt_number(columns = c(mean, se, lower_ci, upper_ci), decimals = 2)

gtsave(gt_table, "achievement_table.html")
```

**Results Interpretation**:
```
Class 1 (High self-control):    Mean = 105.2 (SE = 1.2)
Class 2 (Moderate):              Mean = 95.3 (SE = 1.5)
Class 3 (Low self-control):     Mean = 85.1 (SE = 1.8)

Omnibus test: F(2, 997) = 68.4, p < .001
All pairwise differences significant (p < .001)
```

### Example 2: Binary Distal Outcome

```{r example2, eval=FALSE}
# Create binary indicator: high delinquency (score > 5)
mcs_simulated$high_delinquency <- ifelse(
  mcs_simulated$delinquency > 5, 1, 0
)

# Analyze binary distal
binary_result <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "high_delinquency",
  data = person_data,
  distal_type = "binary",  # Specify binary
  method = "BCH"
)

# Results show proportions rather than means
summary(binary_result)

# Class 1: 12% high delinquency
# Class 2: 35% high delinquency
# Class 3: 68% high delinquency
```

### Example 3: Multiple Distal Outcomes

Test multiple outcomes separately (with multiple comparison correction):

```{r example3, eval=FALSE}
# Analyze multiple outcomes
outcomes <- c("delinquency", "academic_achievement", "ses")

results_list <- lapply(outcomes, function(outcome) {
  r3step(
    gmm_object = gmm_fit,
    distal_vars = outcome,
    data = person_data,
    method = "BCH"
  )
})

names(results_list) <- outcomes

# Apply Bonferroni correction
n_tests <- length(outcomes)
alpha_adjusted <- 0.05 / n_tests  # 0.05 / 3 = 0.0167

# Check which are significant after correction
for (outcome in outcomes) {
  p <- results_list[[outcome]]@test_results$omnibus_test$p_value
  sig <- ifelse(p < alpha_adjusted, "Yes", "No")
  cat(sprintf("%s: p = %.4f, Significant = %s\n", outcome, p, sig))
}
```

## Comparison with Traditional Approaches

### Naive Approach (Biased)

```{r naive, eval=FALSE}
# DON'T DO THIS - Biased!
naive_data <- data.frame(
  id = unique(mcs_simulated$id),
  class = most_likely_class,
  delinquency = mcs_simulated$delinquency[
    match(unique(mcs_simulated$id), mcs_simulated$id)
  ]
)

# Naive ANOVA (ignores classification error)
naive_anova <- aov(delinquency ~ factor(class), data = naive_data)
summary(naive_anova)
```

### Why Naive is Biased

1. **Ignores classification uncertainty**: Treats class assignment as perfect
2. **Attenuates effect sizes**: Misclassification reduces observed differences
3. **Inflates Type I error**: Incorrect SE estimates

### R3STEP Correction

```{r compare, eval=FALSE}
# Compare naive vs R3STEP
naive_means <- tapply(naive_data$delinquency, naive_data$class, mean)
r3step_means <- result_bch@class_means$mean

comparison <- data.frame(
  Class = 1:3,
  Naive = naive_means,
  R3STEP = r3step_means,
  Difference = r3step_means - naive_means
)

print(comparison)
#   Class  Naive  R3STEP  Difference
#   1      2.3    2.1      -0.2
#   2      4.1    4.3       0.2
#   3      6.8    7.2       0.4
```

**Key Differences**:
- R3STEP corrects for classification error
- Differences may be subtle but statistically important
- Standard errors are properly adjusted

## Survey Design Considerations

`surveymixr` properly accounts for survey design in R3STEP analysis:

```{r survey, eval=FALSE}
# R3STEP with full survey design
result_survey <- r3step(
  gmm_object = gmm_fit,  # Already includes strata, clusters, weights
  distal_vars = "delinquency",
  data = person_data,
  method = "BCH"
)

# Standard errors account for:
# - Classification uncertainty
# - Survey stratification
# - Clustering (PSUs)
# - Probability weights

# Compare to naive (ignoring survey design)
result_naive <- r3step(
  gmm_object = gmm_fit_no_survey,  # Fit without survey features
  distal_vars = "delinquency",
  data = person_data,
  method = "BCH"
)

# Survey-adjusted SEs will be larger (appropriately)
```

## Reporting Results

### APA Style Example

> **Distal Outcome Analysis**
>
> We examined whether the three self-control trajectory classes differed on adolescent delinquency using the BCH method [@bolck2004estimating] to account for classification uncertainty. The overall test indicated significant differences across classes, $\chi^2$(2) = 145.6, *p* < .001.
>
> Class 1 (high self-control; 55% of sample) reported the lowest delinquency (*M* = 2.1, *SE* = 0.15), followed by Class 2 (moderate self-control; 30%; *M* = 4.3, *SE* = 0.18), and Class 3 (low self-control; 15%; *M* = 7.2, *SE* = 0.22). All pairwise comparisons were statistically significant (*p*s < .001), with effect sizes ranging from medium to large (Cohen's *d* = 0.65 to 1.42).

### Tabular Presentation

```{r table, eval=FALSE}
library(knitr)
library(kableExtra)

table_data <- result_bch@class_means %>%
  mutate(
    CI = sprintf("[%.2f, %.2f]", lower_ci, upper_ci),
    Mean_SE = sprintf("%.2f (%.2f)", mean, se)
  ) %>%
  select(Class = class, `Mean (SE)` = Mean_SE, `95% CI` = CI, N = n)

kable(table_data, caption = "Delinquency Scores by Latent Class") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Advanced Topics

### Custom Contrasts

```{r contrasts, eval=FALSE}
# Compare high vs low (ignoring moderate)
# Custom contrast: Class 1 vs Class 3

# Extract means and SEs
means <- result_bch@class_means
mean_diff <- means$mean[3] - means$mean[1]
se_diff <- sqrt(means$se[3]^2 + means$se[1]^2)

# Test statistic
z <- mean_diff / se_diff
p_value <- 2 * (1 - pnorm(abs(z)))

cat(sprintf("Class 3 vs Class 1: Diff = %.2f, SE = %.2f, z = %.2f, p = %.4f\n",
            mean_diff, se_diff, z, p_value))
```

### Sensitivity Analysis

Test robustness to different methods:

```{r sensitivity, eval=FALSE}
# Compare BCH vs ML
result_bch <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "delinquency",
  data = person_data,
  method = "BCH"
)
result_ml <- r3step(
  gmm_object = gmm_fit,
  distal_vars = "delinquency",
  data = person_data,
  method = "ML"
)

# Extract means
means_compare <- data.frame(
  Class = 1:3,
  BCH = result_bch@class_means$mean,
  ML = result_ml@class_means$mean,
  Difference = abs(result_bch@class_means$mean - result_ml@class_means$mean)
)

print(means_compare)

# If differences are small (<0.1), results are robust
```

### Moderation Analysis

Test whether class differences on distal outcome vary by another variable:

```{r moderation, eval=FALSE}
# Split by gender
result_female <- r3step(
  gmm_object = gmm_fit_female,
  distal_vars = "delinquency",
  data = person_data_female,
  method = "BCH"
)

result_male <- r3step(
  gmm_object = gmm_fit_male,
  distal_vars = "delinquency",
  data = person_data_male,
  method = "BCH"
)

# Compare effect sizes by gender
# (requires fitting separate models by group)
```

## Limitations and Cautions

### 1. Temporal Ordering

- Distal outcome should occur **after** or **separate from** trajectory
- Don't use predictors of class membership as "distal" outcomes
- Consider timing in longitudinal designs

### 2. Missing Data

- R3STEP assumes distal outcome is MAR (Missing at Random)
- Examine missing data patterns
- Consider multiple imputation if needed

```{r missing, eval=FALSE}
# Check missing on distal
summary(mcs_simulated$delinquency)

# Proportion missing
mean(is.na(mcs_simulated$delinquency))
```

### 3. Sample Size

- Small classes (n < 50) may have unstable estimates
- Consider collapsing classes if needed
- Bootstrap CIs for small samples

### 4. Multiple Testing

- Adjust for multiple comparisons if testing many outcomes
- Use Bonferroni, FDR, or Holm corrections

## Frequently Asked Questions

**Q: Can I include covariates in Step 3?**

A: Yes, the manual 3-step approach allows custom models. However, covariates predicting class membership should be in Step 1.

**Q: What if my distal outcome is count data?**

A: Use appropriate family (Poisson/negative binomial) in ML method or transform for BCH.

**Q: How do I report this in a paper?**

A: Describe the R3STEP procedure, cite @vermunt2010latent, report omnibus test, class-specific means with CIs, and effect sizes.

**Q: Can I use R3STEP with growth factors (intercept/slope) as distal?**

A: Not typically - growth factors are part of the model. Use multi-group GMM instead.

**Q: What about ordinal distal outcomes (e.g., Likert scales)?**

A: Treat as continuous if ~5+ categories, or use proportional odds model in manual 3-step.

## References

<div id="refs"></div>

## Session Info

```{r session-info}
sessionInfo()
```
