---
title: "Real-World Analysis: Millennium Cohort Study Self-Control Trajectories"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Real-World Analysis: Millennium Cohort Study Self-Control Trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

# Introduction

This vignette demonstrates a complete real-world analysis using **surveymixr** with Millennium Cohort Study (MCS) data. We will analyze developmental trajectories of self-control from age 3 to 17, examining how adolescent outcomes (delinquency and academic achievement) vary across trajectory classes.

## Study Context

**Research Question**: Do children follow distinct developmental trajectories of self-control from early childhood to late adolescence, and are these trajectories associated with later adolescent outcomes?

**Data Source**: Millennium Cohort Study (MCS) - UK longitudinal birth cohort
- **Sample**: Approximately 19,000 children born 2000-2002
- **Design**: Complex survey with stratification and clustering
- **Time Points**: Ages 3, 5, 7, 11, 14, and 17 years

## Key Variables

- **Self-Control**: Parent-reported measure of child's ability to control impulses and follow rules
- **Delinquency**: Self-reported antisocial behaviors at age 17
- **Academic Achievement**: Standardized test scores
- **SES**: Socioeconomic status (continuous measure)
- **Survey Design**: Nota Bene: This is simulated data based on real MCS structure

# Load Packages and Data

```{r}
# Load surveymixr and supporting packages
library(surveymixr)
library(ggplot2)
library(knitr)  # For nice tables

data(mcs_simulated)

cat("Dataset dimensions:", dim(mcs_simulated), "\n")
head(mcs_simulated)
```

## Understanding the Data Structure

```{r}
# Data structure
str(mcs_simulated)

# Summary statistics
summary(mcs_simulated[, c("age", "selfcontrol", "delinquency", "academic_achievement", "ses")])
```

# Phase 1: Exploratory Analysis

## Step 1.1: Data Validation

Before fitting models, validate data structure and survey design:

```{r}
validation <- validate_survey_data(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  min_observations = 2,  # Allow at least 2 time points
  verbose = TRUE
)

print(validation)
```

## Step 1.2: Trajectory Visualization

Visualize raw trajectories to understand patterns:

```{r}
# Randomly sample 100 individuals for cleaner spaghetti plot
set.seed(42)
sample_ids <- sample(unique(mcs_simulated$id), 100)
plot_data <- mcs_simulated[mcs_simulated$id %in% sample_ids, ]

# Create simple line plot
library(ggplot2)
ggplot(plot_data, aes(x = age, y = selfcontrol, group = id)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  geom_smooth(aes(group = 1), method = "loess", color = "red", se = TRUE, size = 1) +
  labs(
    title = "Self-Control Trajectories (Random Sample of 100 Individuals)",
    x = "Age (years)",
    y = "Self-Control Score"
  ) +
  theme_minimal()
```

## Step 1.3: Survey Design Summary

Understand the survey design structure:

```{r}
cat("Number of strata:", length(unique(mcs_simulated$stratum)), "\n")
cat("Number of PSUs:", length(unique(mcs_simulated$psu)), "\n")
cat("Sample size:", length(unique(mcs_simulated$id)), "individuals\n")
cat("Individual observations:", nrow(mcs_simulated), "\n")
```

# Phase 2: Model Selection

## Step 2.1: Fit Alternative Class Solutions

We test 1-5 class models using multiple criteria:

```{r}
# Note: Using fewer starts for demonstration; use 500+ for publication
selection_results <- gmm_select(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  classes = 1:5,                    # Test 1-5 class models
  growth_model = "linear",
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  criteria = c("BIC", "aBIC", "entropy"),  # Skip BLRT for speed
  starts = 100,                      # Use 500+ in practice
  cores = 4,
  verbose = TRUE
)

print(selection_results)
```

## Step 2.2: Visualize Model Selection

```{r}
# Plot fit indices comparison
plot_model_selection(
  selection_results,
  criteria = c("bic", "entropy", "aic"),
  show_best = TRUE
)
```

## Step 2.3: Classification Quality Check

```{r}
# Extract and compare classification quality across models
quality_stats <- sapply(
  selection_results@fitted_models,
  function(fit) {
    quality <- classification_quality(fit)
    return(
      c(
        entropy = fit@fit_indices$entropy,
        ave_pp = quality$ave_pp,
        occ = quality$occ
      )
    )
  }
)

# Create comparison table
quality_df <- data.frame(
  n_classes = 1:5,
  t(quality_stats)
)

knitr::kable(quality_df, digits = 3, caption = "Classification Quality by Number of Classes")
```

## Step 2.4: Selection Decision

Based on:
- Minimum BIC at 3 classes
- High entropy (>0.80)
- Theoretical interpretability

```{r}
# Extract recommended classes
recommended <- selection_results@recommended_classes
print(recommended)
```

# Phase 3: Fit Final Model

## Step 3.1: Fit 3-Class Linear Growth Model

```{r}
# Fit final model with more starts (500-1000 recommended)
final_model <- gmm_survey(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  growth_model = "linear",
  n_classes = 3,                    # Based on model selection
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  nest = TRUE,                      # PSUs nested within strata
  starts = 200,                     # Use 500+ for publication
  max_iter = 1000,
  convergence = 1e-6,
  cores = 4,
  keep_data = TRUE,                 # For diagnostics
  verbose = TRUE
)

cat("\nModel fitting complete!\n")
summary(final_model)
```

## Step 3.2: Check Convergence Quality

```{r}
# Diagnose convergence
convergence_diag <- diagnose_convergence(final_model)
print(convergence_diag)

# Visualize log-likelihood distribution
plot(convergence_diag)
```

**Important**: Check that:
- Log-likelihood distribution is unimodal (no multiple local maxima)
- Best solution is well-replicated (>10% of starts)
- Technical error < 5%

## Step 3.3: Visualize Class Trajectories

```{r}
# Plot class trajectories with confidence intervals
plot_trajectories(
  final_model,
  show_ci = TRUE,
  palette = "Set1",
  title = "Three-Class Trajectories of Self-Control (Ages 3-17)"
)
```

**Interpretation**:
- **Class 1**: High starting point, stable trajectory (High Self-Control)
- **Class 2**: Moderate starting point, stable trajectory (Moderate Self-Control)
- **Class 3**: Low starting point, improving trajectory (Low-Improving Self-Control)

## Step 3.4: Classification Quality

```{r}
# Detailed classification diagnostics
cat("Classification Quality:\n")
cat("========================\n\n")

classification <- classification_quality(final_model)
print(classification)

cat("\n\nEntropy:", entropy(final_model), "\n")
cat("(Range: 0-1, >0.80 = good separation)\n")
```

# Phase 4: R3STEP Analysis (Auxiliary Variables)

## Step 4.1: Prepare Person-Level Data

R3STEP requires one row per individual:

```{r}
# Create person-level dataset
person_data <- mcs_simulated[!duplicated(mcs_simulated$id), ]

cat("Person-level data dimensions:", dim(person_data), "\n")
head(person_data[, c("id", "delinquency", "academic_achievement", "ses")])
```

## Step 4.2: Run R3STEP for Delinquency

```{r}
# Analyze delinquency differences across self-control trajectories
delinquency_r3step <- r3step(
  gmm_object = final_model,
  distal_vars = "delinquency",
  data = person_data,
  method = "BCH",        # Bolck-Croon-Hagenaars method
  test_pairwise = TRUE,  # Pairwise comparisons
  weight_distal = TRUE   # Account for survey weights
)

cat("Delinquency R3STEP Results:\n")
cat("============================\n\n")
print(delinquency_r3step@class_means)
```

## Step 4.3: Run R3STEP for Academic Achievement

```{r}
# Analyze academic achievement differences
academic_r3step <- r3step(
  gmm_object = final_model,
  distal_vars = "academic_achievement",
  data = person_data,
  method = "BCH",
  test_pairwise = TRUE,
  weight_distal = TRUE
)

cat("Academic Achievement R3STEP Results:\n")
cat("====================================\n\n")
print(academic_r3step@class_means)
```

## Step 4.4: Statistical Tests and Effect Sizes

```{r}
# Function to display comprehensive R3STEP results
display_r3step_results <- function(r3step_obj, outcome_name) {
  cat("\n\n", outcome_name, "\n", paste(rep("=", nchar(outcome_name)), collapse = ""), "\n\n", sep = "")

  # Omnibus test
  cat("Omnibus Test (Are there overall differences?\n")
  cat("--------------------------------------------\n")
  omni <- r3step_obj@test_results$omnibus_test
  cat(sprintf("Statistic: %.3f\n", omni$statistic))
  cat(sprintf("p-value: %.3f\n", omni$p_value))
  cat(sprintf("Significance: %s\n
", if(omni$p_value < 0.05) "SIGNIFICANT" else "NON-SIGNIFICANT"))

  # Pairwise comparisons
  cat("Pairwise Comparisons:\n")
  cat("---------------------\n")
  pairwise <- r3step_obj@test_results$pairwise_tests
  print(pairwise)
  cat("\n")

  # Effect sizes (Cohen's d)
  cat("Effect Sizes (Cohen's d):\n")
  cat("--------------------------\n")
  cat("Interpretation: 0.2 = small, 0.5 = medium, 0.8 = large\n\n")
  print(r3step_obj@effect_sizes)
}

# Display results
display_r3step_results(delimquency_r3step, "DELINQUENCY")
display_r3step_results(academic_r3step, "ACADEMIC ACHIEVEMENT")
```

**Interpreting Results**:
- **Delinquency**: Often shows that low self-control class has significantly higher delinquency
- **Academic Achievement**: Usually shows that high self-control class has better academic outcomes
- **Effect sizes**: Indicate magnitude of differences (d > 0.5 = practically meaningful)

## Step 4.5: Plot R3STEP Results

```{r}
# Create bar plots showing class differences
par(mfrow = c(1, 2))

# Plot delinquency differences
plot(delimquency_r3step, type = "means",
     main = "Delinquency by Self-Control Class",
     ylab = "Delinquency Score",
     xlab = "Class (based on self-control trajectories)")

# Plot academic achievement differences
plot(academic_r3step, type = "means",
     main = "Academic Achievement by Self-Control Class",
     ylab = "Achievement Score",
     xlab = "Class (based on self-control trajectories)")
```

# Phase 5: Model Diagnostics

## Step 5.1: Influence Diagnostics

Identify influential cases that might affect results:

```{r}
# Check for influential observations
influence_diag <- diagnose_influence(final_model, measure = "all")

# Number of influential cases
influential_count <- length(influence_diag@influential_ids)
cat("Number of influential cases:", influential_count, "\n")
cat("(out of", length(unique(mcs_simulated$id)), "total individuals)\n")

# Summary statistics
print(summary(influence_diag@influence_measures))
```

## Step 5.2: Residual Diagnostics

Check model assumptions:

```{r}
residuals <- residual_diagnostics(final_model)
print(residuals)
```

## Step 5.3: Separation Check

Ensure classes are not perfectly separated:

```{r}
separation <- diagnose_separation(final_model)
print(separation)
```

# Phase 6: Interpretation and Reporting

## Step 6.1: Class Characteristics Summary

```{r}
# Create summary table of class characteristics
cat("CLASS CHARACTERISTICS SUMMARY\n")
cat("===============================\n\n")

# Class sizes and proportions
sizes <- class_proportions(final_model, weighted = TRUE)
for (i in 1:final_model@model_info$n_classes) {
  cat(sprintf("CLASS %d:\n", i))
  cat(sprintf("  Proportion: %.1f%%\n", sizes[i] * 100))

  # Growth parameters
  params <- coef(final_model)[[i]]
  cat(sprintf("  Intercept: %.2f (SE = %.2f)\n",
              params["Intercept"], params["Intercept_SE"]))
  cat(sprintf("  Slope: %.2f (SE = %.2f)\n",
              params["Slope"], params["Slope_SE"]))
  cat("\n")
}
```

## Step 6.2: Theoretical Interpretation

Based on the trajectory shapes and auxiliary variable associations:

**Class 1: High Self-Control (High-Stable)**
- 30-40% of sample (varies by simulation)
- High initial levels maintained across adolescence
- Associated with: Lower delinquency, higher academic achievement
- Characteristics: High impulse control, rule-following

**Class 2: Moderate Self-Control (Moderate-Stable)**
- 40-50% of sample
- Moderate levels maintained across development
- Associated with: Average outcomes across domains
- Characteristics: Typical self-regulation development

**Class 3: Low-Improving Self-Control**
- 20-30% of sample
- Low initial levels but gradual improvement
- Associated with: Higher delinquency, lower academic achievement (but improving)
- Characteristics: Delayed self-regulation development

## Step 6.3: Key Findings

```{r}
cat("KEY FINDINGS:\n")
cat("=============\n\n")

cat("1. TRAJECTORY CLASSES:\n")
cat("   Three distinct trajectories of self-control identified:\n")
cat("   - High-stable (best outcomes)\n")
cat("   - Moderate-stable (typical outcomes)\n")
cat("   - Low-improving (poorer outcomes but catching up)\n\n")

# Get entropy for reporting
cat("2. CLASSIFICATION QUALITY:\n")
cat(sprintf("   Entropy = %.3f (good separation)\n\n", entropy(final_model)))

significance_test <- function(p_value) {
  if (p_value < 0.001) return("***")
  if (p_value < 0.01) return("**")
  if (p_value < 0.05) return("*")
  return("ns")
}

delinquency_p <- delinquency_r3step@test_results$omnibus_test$p_value
academic_p <- academic_r3step@test_results$omnibus_test$p_value

cat("3. AUXILIARY OUTCOMES:\n")
cat(sprintf("   a) Delinquency: Class differences %s (p = %.3f)\n",
            significance_test(delinquent_p), delinquency_p))
cat(sprintf("   b) Academic Achievement: Class differences %s (p = %.3f)\n",
            significance_test(academic_p), academic_p))
```

# Phase 7: Publication-Ready Tables and Figures

## Step 7.1: Parameter Estimates Table

```{r}
# Extract parameter estimates for publication table
extract_parameters <- function(fit) {
  all_coefs <- coef(fit)
  params <- do.call(rbind, lapply(all_coefs, function(x) {
    data.frame(
      Class = x["Class"],
      Intercept = x["Intercept"],
      Intercept_SE = x["Intercept_SE"],
      Slope = x["Slope"],
      Slope_SE = x["Slope_SE"],
      Proportion = x["Class_Proportion"]
    )
  }))
  return(params)
}

# Create publication-ready table
par_table <- extract_parameters(final_model)
knitr::kable(
  par_table,
  digits = 3,
  caption = "Growth Parameters by Self-Control Trajectory Class"
)
```

## Step 7.2: Fit Indices Table

```{r}
# Create fit indices table
fit_table <- data.frame(
  Model = paste0(1:5, "-Class"),
  LogLik = sapply(selection_results@fitted_models, logLik),
  AIC = sapply(selection_results@fitted_models, AIC),
  BIC = sapply(selection_results@fitted_models, BIC),
  Entropy = sapply(selection_results@fitted_models, function(x) x@fit_indices$entropy)
)

knitr::kable(
  fit_table,
  digits = 2,
  caption = "Model Fit Statistics for Alternative Class Solutions"
)
```

## Step 7.3: R3STEP Results Table

```{r}
# Create comprehensive R3STEP results table
extract_r3step_all <- function(r3step_obj) {
  means <- r3step_obj@class_means
  pairwise <- r3step_obj@test_results$pairwise_tests
  effects <- r3step_obj@effect_sizes

  # Combine into comprehensive table
  results <- data.frame(
    Class = 1:nrow(means),
    Mean = means$mean,
    SE = means$se,
    Comparison = paste0("vs ", effects$comparison_class),
    Effect_Size = effects$cohens_d,
    Effect_Interpretation = ifelse(
      abs(effects$cohens_d) < 0.5, "Small",
      ifelse(abs(effects$cohens_d) < 0.8, "Medium", "Large")
    )
  )
  return(results)
}

cat("DELINQUENCY R3STEP RESULTS:\n")
knitr::kable(
  extract_r3step_all(delinquent_r3step),
  digits = 3,
  caption = "Delinquency Differences Across Self-Control Classes"
)

cat("\n\nACADEMIC ACHIEVEMENT R3STEP RESULTS:\n")
knitr::kable(
  extract_r3step_all(academic_r3step),
  digits = 3,
  caption = "Academic Achievement Differences Across Self-Control Classes"
)
```

# Summary and Best Practices

## Key Takeaways

1. **Always validate your data** before fitting models
2. **Use multiple criteria** for model selection (never rely on just one!)
3. **Check convergence thoroughly** - inadequate starts can miss true solution
4. **Validate classification quality** with entropy, AvePP, and other metrics
5. **Use R3STEP for auxiliary variables** to avoid classification bias
6. **Run comprehensive diagnostics** before interpreting results

## Recommended Workflow

For your own research:

1. **Data Preparation**
   - Validate data structure
   - Check survey design variables
   - Examine missing patterns

2. **Exploratory Analysis**
   - Visualize trajectories
   - Estimate 2-5 class models

3. **Model Selection**
   - Compare information criteria
   - Check classification quality
   - Validate with BLRT

4. **Final Model**
   - Fit selected model with many starts
   - Check convergence
   - Thorough diagnostics

5. **Auxiliary Analysis**
   - R3STEP for distal outcomes
   - R3STEP for predictors
   - Compare across classes

6. **Reporting**
   - Document all decisions
   - Include diagnostics
   - Archive analysis scripts

## Next Steps

- **Cross-validation** to assess stability of classes
- **Sensitivity analyses** with different growth models
- **Multiple imputation** if missing data concerns
- **Subgroup analyses** (e.g., by gender, SES)
- **Longitudinal prediction** of later adult outcomes

## Citing This Analysis

If you use this analysis in your research:

```{r}
citation("surveymixr")
```

---

**End of Vignette**

*This vignette demonstrates a complete real-world application of surveymixr. For questions, visit: https://github.com/siyangni/surveymixr*
