---
title: "Technical Details and Algorithms"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical Details and Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(surveymixr)
```

## Introduction

This vignette provides technical details on the statistical methods and algorithms implemented in `surveymixr`. It covers:

- Growth mixture modeling framework
- Complex survey design integration
- EM algorithm implementation
- Standard error computation
- Bootstrap likelihood ratio test (BLRT)
- Model fit indices

For an introduction to using the package, see `vignette("surveymixr-intro")`.

## Statistical Framework

### Growth Mixture Models

Growth mixture models (GMMs) combine growth curve modeling with latent class analysis to identify distinct subgroups (classes) with different developmental trajectories [@muthen2000finite].

#### Individual-Level Model

For individual $i$ in class $k$ at time $t$:

$$y_{it} | k \sim N(\mu_{itk}, \sigma^2_k)$$

where the mean trajectory is:

$$\mu_{itk} = \beta_{0k} + \beta_{1k} \cdot time_{it} + \beta_{2k} \cdot time_{it}^2 + \cdots$$

#### Class Membership

The probability that individual $i$ belongs to class $k$:

$$P(C_i = k | \mathbf{X}_i) = \frac{\exp(\gamma_{k0} + \gamma_k^T \mathbf{X}_i)}{\sum_{j=1}^K \exp(\gamma_{j0} + \gamma_j^T \mathbf{X}_i)}$$

where $\mathbf{X}_i$ are time-invariant covariates (optional).

### Complex Survey Design Features

Survey data often involve:

1. **Stratification**: Population divided into strata, samples drawn from each
2. **Clustering**: Multi-stage sampling (e.g., schools, then students)
3. **Unequal probability weights**: Some units more likely to be sampled

#### Survey-Weighted Likelihood

The survey-weighted log-likelihood is:

$$\ell_w(\theta) = \sum_{i=1}^n w_i \log f(y_i | \theta)$$

where $w_i$ is the survey weight for individual $i$.

**Key innovation**: `surveymixr` integrates survey weights directly into the EM algorithm, unlike post-hoc weighting approaches.

### Integration: Survey-Weighted GMM

Combining GMM with survey design:

$$\ell_w(\theta) = \sum_{i=1}^n w_i \log \left[ \sum_{k=1}^K \pi_k \prod_{t=1}^{T_i} f(y_{it} | \theta_k) \right]$$

where:
- $w_i$ = survey weight
- $\pi_k$ = class proportion (weighted)
- $\theta_k$ = class-specific parameters
- $T_i$ = number of observations for individual $i$

## Estimation Algorithm

### EM Algorithm Overview

The Expectation-Maximization (EM) algorithm [@dempster1977maximum] iterates between:

1. **E-step**: Compute posterior class probabilities
2. **M-step**: Update parameters given posterior probabilities

#### E-Step: Posterior Probabilities

For individual $i$ and class $k$:

$$\tau_{ik} = \frac{w_i \pi_k \prod_{t=1}^{T_i} f(y_{it} | \theta_k)}{\sum_{j=1}^K w_i \pi_j \prod_{t=1}^{T_i} f(y_{it} | \theta_j)}$$

where $\tau_{ik} = P(C_i = k | y_i, \theta)$ is the posterior probability.

#### M-Step: Parameter Updates

**Class proportions** (weighted):

$$\pi_k = \frac{\sum_{i=1}^n w_i \tau_{ik}}{\sum_{i=1}^n w_i}$$

**Growth parameters** (class $k$):

Solve weighted regression:

$$\hat{\beta}_k = \arg\min_{\beta_k} \sum_{i=1}^n \sum_{t=1}^{T_i} w_i \tau_{ik} (y_{it} - \mathbf{X}_{it}^T \beta_k)^2$$

**Variance** (class $k$):

$$\hat{\sigma}^2_k = \frac{\sum_{i=1}^n \sum_{t=1}^{T_i} w_i \tau_{ik} (y_{it} - \hat{y}_{it})^2}{\sum_{i=1}^n \sum_{t=1}^{T_i} w_i \tau_{ik}}$$

#### Convergence Criterion

Stop when:

$$|\ell_w^{(t+1)} - \ell_w^{(t)}| < \epsilon$$

Default: $\epsilon = 10^{-6}$

### Missing Data Handling (FIML)

`surveymixr` uses Full Information Maximum Likelihood (FIML) for missing data under the Missing at Random (MAR) assumption.

For individual $i$ with observed data $y_i^{obs}$:

$$\ell_w(\theta | y_i^{obs}) = \sum_{k=1}^K w_i \tau_{ik} \log \prod_{t \in obs_i} f(y_{it} | \theta_k)$$

**Advantages**:
- Uses all available data
- Unbiased under MAR
- No need for imputation

### Multiple Random Starts

Local maxima are common in mixture models. `surveymixr` uses multiple random starts:

1. Generate `starts` (default 500) random initial parameter values
2. Run EM from each starting point (parallelized)
3. Select solution with highest log-likelihood
4. Check for local maxima with `diagnose_convergence()`

**Example**:

```{r random-starts, eval=FALSE}
fit <- gmm_survey(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  n_classes = 3,
  starts = 500,  # Multiple random starts
  cores = 4      # Parallel processing
)

# Check for local maxima
diagnose_convergence(fit)
```

## Standard Errors

### Sandwich Estimator

`surveymixr` uses the sandwich (robust) variance estimator to account for:
- Model misspecification
- Clustering
- Stratification

$$\mathbf{V}(\hat{\theta}) = \mathbf{H}^{-1} \mathbf{J} \mathbf{H}^{-1}$$

where:
- $\mathbf{H}$ = Hessian (observed information matrix)
- $\mathbf{J}$ = Outer product of score contributions (meat)

### Accounting for Survey Design

**Clustering** adjustment:

$$\mathbf{J} = \sum_{h=1}^H \sum_{c=1}^{C_h} \left(\sum_{i \in c} \mathbf{s}_i\right) \left(\sum_{i \in c} \mathbf{s}_i\right)^T$$

where:
- $H$ = number of strata
- $C_h$ = number of clusters in stratum $h$
- $\mathbf{s}_i$ = score contribution for individual $i$

**Effect**: Standard errors increase to reflect within-cluster correlation (design effect > 1).

### Numerical Derivatives

Since analytical derivatives are complex, `surveymixr` uses numerical derivatives via the `numDeriv` package:

- **Hessian**: Richardson extrapolation
- **Gradient**: Central differences

**Confidence intervals**:

$$\hat{\theta} \pm 1.96 \times SE(\hat{\theta})$$

## Bootstrap Likelihood Ratio Test (BLRT)

The BLRT [@mclachlan1987mixture] tests whether a $k$-class model fits better than a $(k-1)$-class model.

### Algorithm

1. Fit $(k-1)$-class model, obtain $\hat{\theta}_{k-1}$
2. Fit $k$-class model, obtain $\hat{\theta}_k$
3. Compute observed LRT statistic:
   $$LRT_{obs} = 2(\ell(\hat{\theta}_k) - \ell(\hat{\theta}_{k-1}))$$
4. **Bootstrap** (repeat $B$ times):
   - Simulate data from $(k-1)$-class model
   - Fit both $(k-1)$ and $k$-class models
   - Compute $LRT_b$
5. $p$-value = proportion of $LRT_b > LRT_{obs}$

**Interpretation**:
- $p < 0.05$: Reject $(k-1)$ classes, favor $k$ classes
- $p \geq 0.05$: No evidence for $k$ classes

**Example**:

```{r blrt, eval=FALSE}
selection <- gmm_select(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  min_classes = 1,
  max_classes = 5,
  run_blrt = TRUE,
  blrt_samples = 500,  # Bootstrap replicates
  cores = 4
)

# View BLRT results
summary(selection)
```

## Model Fit Indices

### Information Criteria

**AIC** (Akaike Information Criterion):

$$AIC = -2\ell(\hat{\theta}) + 2p$$

**BIC** (Bayesian Information Criterion):

$$BIC = -2\ell(\hat{\theta}) + p \log(n)$$

**Sample-size adjusted BIC**:

$$aBIC = -2\ell(\hat{\theta}) + p \log\left(\frac{n+2}{24}\right)$$

where:
- $\ell(\hat{\theta})$ = maximized log-likelihood
- $p$ = number of parameters
- $n$ = sample size (effective sample size for weighted data)

**Lower values indicate better fit**.

### Entropy

Entropy measures classification certainty [@ramaswamy1993empirical]:

$$E = 1 - \frac{\sum_{i=1}^n \sum_{k=1}^K \tau_{ik} \log(\tau_{ik})}{n \log(K)}$$

- Range: $[0, 1]$
- Higher values = better classification
- $E > 0.8$ is considered good
- $E > 0.9$ is excellent

### Average Posterior Probability (AvePP)

For each class $k$:

$$AvePP_k = \frac{\sum_{i: \arg\max_j \tau_{ij} = k} \tau_{ik}}{n_k}$$

where $n_k$ = number of individuals assigned to class $k$.

**Interpretation**:
- AvePP > 0.7 for all classes is acceptable
- AvePP > 0.8 is good

### Odds of Correct Classification (OCC)

$$OCC_k = \frac{AvePP_k / (1 - AvePP_k)}{\pi_k / (1 - \pi_k)}$$

where $\pi_k$ = class proportion.

- OCC > 5 indicates good classification

## Computational Efficiency

### Parallelization

Random starts are parallelized using the `parallel` package:

```{r parallel, eval=FALSE}
# Detect cores
n_cores <- parallel::detectCores() - 1

fit <- gmm_survey(
  ...,
  cores = n_cores
)
```

### Memory Management

For large datasets:
- Use sparse matrices where applicable
- Process data in chunks for FIML
- Limit random starts for initial exploration

### Typical Run Times

On a modern laptop (4 cores):

| n | T | K | starts | Time |
|---|---|---|--------|------|
| 1,000 | 4 | 2 | 100 | ~1 min |
| 1,000 | 4 | 3 | 500 | ~5 min |
| 5,000 | 6 | 3 | 500 | ~15 min |
| 10,000 | 6 | 4 | 1000 | ~45 min |

**Tips for speed**:
1. Start with fewer random starts (50-100) for exploration
2. Use parallel processing (4+ cores)
3. Reduce to final model with more starts (500-1000)

## Comparison with Mplus

`surveymixr` produces results consistent with Mplus:

| Feature | surveymixr | Mplus |
|---------|-----------|--------|
| Parameter estimates | ✓ | ✓ |
| Standard errors | ✓ (sandwich) | ✓ |
| Survey weights | ✓ | ✓ |
| BLRT | ✓ | ✓ |
| Speed | Comparable | Comparable |
| Cost | Free | $\$\$\$$ |
| R integration | Native | Via MplusAutomation |

See `vignette("mplus-validation")` for detailed comparisons.

## References

<div id="refs"></div>

## Session Info

```{r session-info}
sessionInfo()
```
