---
title: "surveymixr User Manual"
author: "Siyang Ni"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{surveymixr User Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

# Introduction

## About This Manual

This comprehensive manual covers all aspects of the **surveymixr** package, from basic concepts to advanced applications. It consolidates information from all package vignettes into a single, unified guide suitable for both beginners and advanced users.

## What is surveymixr?

**surveymixr** is the first R package to implement growth mixture modeling (GMM) with full integration of complex survey design features. It enables rigorous analysis of longitudinal data from large-scale surveys (e.g., Millennium Cohort Study, Add Health, NLSY, ECLS) that were previously only possible in proprietary software like Mplus.

### Key Features

- **Complex Survey Integration**: Full support for stratification, clustering (PSUs), and probability weights
- **Survey-Adjusted Standard Errors**: Sandwich estimators accounting for design effects
- **Multiple Outcome Types**: Continuous, binary, ordinal, and count outcomes
- **Advanced Model Selection**: Bootstrap LRT (BLRT), Lo-Mendell-Rubin test, cross-validation
- **R3STEP Method**: Classification uncertainty correction for auxiliary variables
- **Random Effects**: Individual-level variance in growth parameters
- **Time-Varying Covariates**: Within-person and between-person effects
- **Comprehensive Diagnostics**: Convergence tracking, influence detection, separation checks
- **Professional Visualization**: Publication-ready plots with ggplot2 and plotly
- **Parallel Processing**: Efficient handling of 500-1000+ random starts

### Why Choose surveymixr?

**Comparison with Existing R Packages:**

| Feature | surveymixr | lcmm | flexmix | lavaan | openmx |
|---------|------------|------|---------|--------|---------|
| Growth mixture models | ✓ | ✓ | Partial | ✗ | ✗ |
| Complex survey design | ✓ | ✗ | ✗ | Partial | ✗ |
| Stratification + clustering | ✓ | ✗ | ✗ | ✗ | ✗ |
| Probability weights | ✓ | ✗ | ✗ | ✗ | ✗ |
| Survey-adjusted SEs | ✓ | ✗ | ✗ | ✗ | ✗ |
| BLRT | ✓ | ✗ | ✗ | ✗ | ✗ |
| R3STEP | ✓ | ✗ | ✗ | ✗ | ✗ |
| 1000+ random starts | ✓ | ✗ | ✗ | ✗ | ✗ |

---

# Part I: Getting Started

## Installation

### From CRAN (Recommended)

```{r}
install.packages("surveymixr")
```

### Development Version from GitHub

```{r}
# Install devtools if needed
install.packages("devtools")

# Install surveymixr
devtools::install_github("siyangni/surveymixr")
```

### System Requirements

- R (>= 4.0.0)
- No compiled code (pure R package, works on all platforms)

## Loading the Package

```{r}
library(surveymixr)

# Check version
packageVersion("surveymixr")
```

## Quick Start Example

### Load Example Data

```{r}
# Load simulated Millennium Cohort Study data
data(mcs_simulated)

# Examine data structure
head(mcs_simulated)
str(mcs_simulated)
```

The data contains:
- `id`: Individual identifier
- `age`: Time points (3, 5, 7, 11, 14, 17 years)
- `selfcontrol`: Outcome variable (self-control scores)
- `stratum`, `psu`, `weight`: Survey design variables
- `true_class`: True class membership (for validation)

### Fit a Basic 3-Class Model

```{r}
fit <- gmm_survey(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  n_classes = 3,
  growth_model = "linear",
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  starts = 100,
  cores = 4,
  verbose = TRUE
)
```

### View Results

```{r}
# Comprehensive summary
summary(fit)

# Extract fit indices
extract_fit_indices(fit)

# Classification quality
classification_quality(fit)
```

### Visualize Trajectories

```{r}
# Static plot
plot(fit, type = "trajectories")

# Interactive plot
plot_interactive(fit, type = "trajectories")
```

---

# Part II: Core Concepts

## Growth Mixture Models

### What are GMMs?

Growth mixture models (GMMs) identify subgroups with distinct developmental trajectories in longitudinal data. They combine:

1. **Growth Models**: Model change over time within individuals
2. **Mixture Models**: Identify latent subgroups (classes) with different patterns
3. **Survey Design**: Account for complex sampling structures

### When to Use GMMs

Use GMMs when:
- You have longitudinal data (3+ time points)
- You suspect heterogeneity in trajectories
- You want to identify distinct developmental patterns
- Data comes from complex surveys with sampling weights

### Model Components

```
Y_it = f(time) + ε_it        # Within-class growth model
P(Class = k) = π_k           # Class membership probabilities
```

Where:
- `Y_it`: Outcome for individual i at time t
- `f(time)`: Growth function (linear, quadratic, etc.)
- `ε_it`: Residual error
- `π_k`: Proportion in class k

## Complex Survey Design

### Why Survey Design Matters

Ignoring survey design features leads to:
- **Biased standard errors** (usually underestimated)
- **Invalid p-values and confidence intervals**
- **Incorrect inference** about class differences

### Survey Features in surveymixr

**1. Stratification**
- Divides population into non-overlapping groups
- Example: Geographic regions, urban/rural

**2. Clustering (PSUs)**
- Individuals sampled in groups
- Example: Schools, neighborhoods
- Creates within-cluster correlation

**3. Probability Weights**
- Adjust for unequal sampling probabilities
- Example: Oversampling of minorities
- Ensures nationally representative estimates

**4. Survey-Adjusted Standard Errors**
- Uses sandwich estimator
- Accounts for design effects
- Produces valid inference

### Specifying Survey Design

```{r}
fit <- gmm_survey(
  data = your_data,
  id = "id",
  time = "time",
  outcome = "outcome",
  n_classes = 3,

  # Survey design features:
  strata = "stratum_var",     # Stratification variable
  cluster = "psu_var",        # Primary sampling unit
  weights = "weight_var",     # Sampling weights
  nest = TRUE                 # PSUs nested within strata
)
```

---

# Part III: Basic Model Fitting

## Model Specification

### Required Arguments

```{r}
fit <- gmm_survey(
  data = your_data,           # Long-format data frame
  id = "id_variable",         # Individual ID
  time = "time_variable",     # Time variable
  outcome = "outcome_var",    # Outcome variable
  n_classes = 3               # Number of latent classes
)
```

### Growth Model Types

**1. Linear Growth (Default)**
```{r}
fit <- gmm_survey(..., growth_model = "linear")
# Y_it = β0 + β1*time + ε
```

**2. Quadratic Growth**
```{r}
fit <- gmm_survey(..., growth_model = "quadratic")
# Y_it = β0 + β1*time + β2*time² + ε
```

**3. Nonlinear Growth**
```{r}
fit <- gmm_survey(..., growth_model = "nonlinear")
# Flexible spline-based trajectories
```

**4. Free Basis (Time Scores)**
```{r}
fit <- gmm_survey(..., growth_model = "free_basis")
# Estimates time loadings freely
```

### Estimation Options

```{r}
fit <- gmm_survey(
  ...,
  starts = 500,              # Number of random starts (higher = better)
  convergence = 1e-6,        # Convergence criterion
  max_iter = 1000,           # Maximum EM iterations
  cores = 4,                 # Parallel processing cores
  keep_data = TRUE,          # Store data in object (for diagnostics)
  verbose = TRUE             # Print progress
)
```

**Recommended Settings:**
- `starts`: 200-500 for exploration, 500-1000 for final models
- `cores`: Set to number of available CPU cores
- `convergence`: Default (1e-6) is usually adequate

## Examining Results

### Summary Output

```{r}
summary(fit)
```

Shows:
- Model information (classes, growth model, sample size)
- Survey design features
- Fit indices (AIC, BIC, aBIC, entropy, log-likelihood)
- Class proportions (weighted and unweighted)
- Growth parameters by class
- Standard errors

### Extracting Components

```{r}
# Coefficients
coef(fit)

# Variance-covariance matrix
vcov(fit)

# Log-likelihood
logLik(fit)

# Information criteria
AIC(fit)
BIC(fit)

# Fit indices table
extract_fit_indices(fit)

# Predicted trajectories
trajectories <- extract_trajectories(fit, class = 1)
```

### Classification

```{r}
# Posterior probabilities
post_probs <- fit@posterior_probs
head(post_probs)

# Most likely class assignments
class_assignments <- fit@class_assignments
table(class_assignments)

# Classification quality
classification_quality(fit)

# Entropy (higher = better separation)
entropy(fit)

# Class proportions
class_proportions(fit)
```

---

# Part IV: Model Selection

## Determining the Number of Classes

Model selection is **critical** in GMM. Use multiple criteria:

### 1. Information Criteria

**Lower is better** for IC measures:

```{r}
# Compare models with different class counts
selection <- gmm_select(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  classes = 1:5,              # Test 1-5 classes
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  criteria = c("BIC", "aBIC", "entropy"),
  starts = 200,
  cores = 4
)

# View results
print(selection)
plot_model_selection(selection)
```

**Criteria:**
- **BIC**: Bayesian Information Criterion (most commonly used)
- **aBIC**: Sample-size adjusted BIC (for large samples)
- **AIC**: Akaike Information Criterion (less penalized)

**Decision Rule:**
- Choose model where IC reaches minimum or "elbow"
- Lower values indicate better fit
- Balance fit vs. parsimony

### 2. Bootstrap Likelihood Ratio Test (BLRT)

**Gold standard** for class enumeration:

```{r}
selection <- gmm_select(
  ...,
  criteria = c("BIC", "BLRT"),
  blrt_samples = 100,        # Bootstrap samples
  blrt_cores = 4             # Parallel processing
)
```

**Interpretation:**
- Tests k-class vs. (k-1)-class model
- Significant p-value → k-class model is better
- Stop when p-value becomes non-significant

### 3. Lo-Mendell-Rubin (LMR) Test

**Faster alternative** to BLRT (no bootstrap):

```{r}
# Fit two models
fit2 <- gmm_survey(..., n_classes = 2, ...)
fit3 <- gmm_survey(..., n_classes = 3, ...)

# LMR test
lmr_result <- lmr_test(fit3, fit2, adjusted = TRUE)
print(lmr_result)
```

**Advantages:**
- Much faster than BLRT
- Good approximation in large samples

**Disadvantages:**
- Less accurate in small samples
- May be anti-conservative

### 4. Cross-Validation

**Predictive performance** approach:

```{r}
cv_results <- gmm_cv(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  classes = 1:5,
  k_folds = 5,              # 5-fold CV
  stratified = TRUE,         # Maintain class proportions
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  starts = 100,
  cores = 4
)

print(cv_results)
plot(cv_results)
```

**Metrics:**
- Predictive log-likelihood
- Mean squared prediction error (MSPE)
- One-standard-error rule

### 5. Classification Quality

```{r}
# For each candidate model
classification_quality(fit)
```

**Metrics:**
- **Entropy** (0-1): Higher = better separation (>0.80 good)
- **AvePP** (Average Posterior Probability): >0.70 adequate, >0.80 good
- **OCC** (Odds of Correct Classification): >5 adequate

### Recommended Strategy

1. **Start with IC** (BIC, aBIC) to narrow range
2. **Use BLRT** to confirm optimal k
3. **Check classification quality** (entropy, AvePP)
4. **Consider interpretability** and theoretical fit
5. **Validate with CV** if computational resources allow

**Example Workflow:**

```{r}
# Step 1: Comprehensive selection
selection <- gmm_select(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  classes = 1:6,
  strata = "stratum", cluster = "psu", weights = "weight",
  criteria = c("BIC", "aBIC", "entropy", "BLRT"),
  blrt_samples = 100,
  starts = 500,
  cores = 4
)

print(selection)
plot_model_selection(selection)

# Step 2: Fit final model
best_k <- selection@recommended_classes$BLRT
fit_final <- gmm_survey(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  n_classes = best_k,
  strata = "stratum", cluster = "psu", weights = "weight",
  starts = 1000,  # More starts for final model
  cores = 4
)

# Step 3: Validate
diagnose_convergence(fit_final)
classification_quality(fit_final)
```

---

# Part V: Outcome Types

## Continuous Outcomes (Default)

```{r}
fit <- gmm_survey(
  data = your_data,
  id = "id", time = "time", outcome = "continuous_var",
  n_classes = 3,
  outcome_type = "continuous"  # Default
)
```

**Assumptions:**
- Normal distribution
- Linear relationship (or specify quadratic/nonlinear)

## Binary Outcomes

```{r}
fit_binary <- gmm_survey_binary(
  data = your_data,
  id = "id", time = "time", outcome = "binary_var",  # 0/1
  n_classes = 3,
  link = "probit",           # or "logit"
  strata = "stratum",
  cluster = "psu",
  weights = "weight"
)
```

**Link Functions:**
- `"probit"` (default): Assumes normal latent variable
- `"logit"`: Logistic model

**Example:** Smoking initiation trajectories (yes/no)

## Ordinal Outcomes

```{r}
fit_ordinal <- gmm_survey_ordinal(
  data = your_data,
  id = "id", time = "time", outcome = "ordinal_var",  # 1,2,3,4,5
  n_classes = 3,
  strata = "stratum",
  cluster = "psu",
  weights = "weight"
)
```

**Model:**
- Proportional odds model
- Ordered categories (e.g., Likert scales)

**Example:** Health rating trajectories (Poor, Fair, Good, Very Good, Excellent)

## Count Outcomes

```{r}
fit_count <- gmm_survey_count(
  data = your_data,
  id = "id", time = "time", outcome = "count_var",  # 0,1,2,3,...
  n_classes = 3,
  distribution = "poisson",  # or "negbin", "zip"
  strata = "stratum",
  cluster = "psu",
  weights = "weight"
)
```

**Distributions:**
- `"poisson"`: Simple count model
- `"negbin"`: Negative binomial (handles overdispersion)
- `"zip"`: Zero-inflated Poisson (excess zeros)

**Example:** Delinquency trajectories (number of incidents)

---

# Part VI: Advanced Features

## Random Effects Models

### Why Random Effects?

Standard GMM assumes **no within-class variance** in growth parameters. Random effects add:

- Individual-level deviations from class means
- More realistic modeling of heterogeneity
- Better fit for highly variable trajectories

### Fitting RE Models

```{r}
fit_re <- gmm_survey_re(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  n_classes = 3,
  random = ~ 1 + age,        # Random intercept and slope
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  starts = 200,
  cores = 4
)
```

### Extracting Random Effects

```{r}
# Get BLUPs (Best Linear Unbiased Predictors)
random_effects <- get_random_effects(fit_re)
head(random_effects)

# Variance components
variance_components(fit_re)

# Intraclass correlations
calculate_icc(fit_re)

# Visualize
plot_random_effects(fit_re, type = "density")
plot_random_effects(fit_re, type = "caterpillar")
```

## Time-Varying Covariates

### Why TVCs?

Model how **within-person changes** in predictors affect outcomes.

**Example:** How does stress (changing over time) affect mental health trajectories?

### Fitting TVC Models

```{r}
fit_tvc <- gmm_survey_tvc(
  data = your_data,
  id = "id", time = "time", outcome = "outcome",
  n_classes = 3,
  tvc_formula = ~ stress + support,  # Time-varying predictors
  tvc_effect = "both",               # Within + between effects
  lag = 1,                           # Lagged effects
  strata = "stratum",
  cluster = "psu",
  weights = "weight"
)
```

**Effect Types:**
- `"direct"`: Contemporaneous effects
- `"indirect"`: Mediated through growth
- `"both"`: Within-person + between-person decomposition
- `"interaction"`: TVC × time interaction

### Interpreting TVC Results

```{r}
# Extract effects
tvc_effects <- extract_tvc_effects(fit_tvc)
print(tvc_effects)

# Statistical tests
test_tvc_effects(fit_tvc)

# Visualize
plot_tvc_effects(fit_tvc, covariate = "stress")
```

---

# Part VII: Auxiliary Variable Analysis

## The R3STEP Approach

### Why R3STEP?

**Problem:** Direct inclusion of auxiliary variables (e.g., demographics, outcomes) in GMM can:
- Alter class formation
- Lead to biased estimates
- Change class membership

**Solution:** R3STEP performs auxiliary analysis **after** class formation, accounting for classification uncertainty.

### When to Use R3STEP

Use R3STEP to:
- Compare classes on **distal outcomes** (e.g., academic achievement)
- Examine **demographic differences** between classes
- Test **predictors** of class membership
- Avoid **bias from classification error**

### Fitting R3STEP Models

```{r}
# Step 1: Fit GMM (without auxiliary variables)
fit <- gmm_survey(
  data = mcs_simulated,
  id = "id", time = "age", outcome = "selfcontrol",
  n_classes = 3,
  strata = "stratum", cluster = "psu", weights = "weight",
  starts = 500
)

# Step 2: Prepare distal variables (one row per individual)
distal_data <- mcs_simulated[!duplicated(mcs_simulated$id), ]

# Step 3: R3STEP analysis
r3step_results <- r3step(
  gmm_object = fit,
  distal_vars = c("academic_achievement", "delinquency", "ses"),
  data = distal_data,
  method = "BCH"             # or "ML", "manual"
)
```

### Methods

**1. BCH (Bolck-Croon-Hagenaars)**
- Default method
- Most commonly used
- Good balance of bias and efficiency

**2. ML (Maximum Likelihood)**
- More efficient
- Slightly more complex

**3. Manual (Three-Step)**
- Simple but potentially biased
- Use for comparison only

### Interpreting R3STEP Results

```{r}
# Comprehensive summary
summary(r3step_results)

# Class-specific means for distal variables
print(r3step_results@class_means)

# Omnibus tests (overall differences across classes)
print(r3step_results@test_results)

# Effect sizes (Cohen's d for pairwise comparisons)
print(r3step_results@effect_sizes)
```

**Interpretation:**
- Omnibus test: Are there overall differences?
- Pairwise tests: Which classes differ?
- Effect sizes: How large are differences? (d > 0.5 = medium, d > 0.8 = large)

---

# Part VIII: Diagnostics

## Convergence Diagnostics

### Why Check Convergence?

GMM estimation can find **local maxima**. Multiple random starts help find the global maximum.

### Diagnose Convergence

```{r}
conv_diag <- diagnose_convergence(fit)
print(conv_diag)

# Visualize distribution of log-likelihoods
plot(conv_diag)
```

**Red Flags:**
- ⚠️ Best solution replicated <10% of the time
- ⚠️ Multiple distinct local maxima
- ⚠️ Wide range of log-likelihood values

**Solutions:**
- Increase `starts` (e.g., 500 → 1000)
- Try different initialization methods
- Consider simpler model (fewer classes)

## Influence Diagnostics

### Detecting Influential Cases

```{r}
# Requires keep_data = TRUE in original fit
fit <- gmm_survey(..., keep_data = TRUE)

# Influence diagnostics
influence <- diagnose_influence(fit, measure = "all")
print(influence)

# Get influential IDs
influential_ids <- influence@influential_ids

# Examine influential cases
mcs_simulated[mcs_simulated$id %in% influential_ids, ]
```

**Measures:**
- **Cook's Distance**: Overall influence on parameters
- **DFBETAS**: Parameter-specific influence
- **Leverage**: Unusual predictor patterns

**Actions:**
- Examine influential cases for data quality issues
- Sensitivity analysis: Refit without influential cases
- Consider robustness checks

## Separation Diagnostics

### Detecting Perfect Separation

```{r}
sep_diag <- diagnose_separation(fit)
print(sep_diag)
```

**Problems:**
- Perfect separation: Classes completely distinct
- Quasi-separation: Near-perfect distinction
- Empty classes: Class with 0 or very few members

**Solutions:**
- Reduce number of classes
- Check for data issues or outliers
- Add constraints

## Residual Diagnostics

```{r}
resid_diag <- residual_diagnostics(fit)
print(resid_diag)
```

**Checks:**
- Normality (Shapiro-Wilk test, Q-Q plots)
- Outliers
- Autocorrelation
- Class-specific patterns

---

# Part IX: Visualization

## Static Plots (ggplot2)

### Trajectory Plots

```{r}
# Class-specific mean trajectories
plot(fit, type = "trajectories")

# With confidence intervals
plot_trajectories(fit, show_ci = TRUE)

# Customize
plot_trajectories(fit,
  show_ci = TRUE,
  color_palette = c("red", "blue", "green")
)
```

### Model Selection Plots

```{r}
# Compare fit indices
plot_model_selection(selection)

# Specific criteria
plot_model_selection(selection, criteria = c("BIC", "entropy"))
```

### Classification Plots

```{r}
# Posterior probability distributions
plot_posterior_dist(fit)

# Class size comparison
plot_class_comparison(fit)
```

## Interactive Plots (plotly)

```{r}
# Interactive trajectories (hover, zoom, pan)
plot_interactive(fit, type = "trajectories")

# Spaghetti plot (individual trajectories)
plot_interactive(fit, type = "spaghetti", n_individuals = 50)

# Save to HTML
library(htmlwidgets)
p <- plot_interactive(fit)
saveWidget(p, "trajectories.html")
```

## Specialized Plots

```{r}
# Random effects
plot_random_effects(fit_re, type = "density")
plot_random_effects(fit_re, type = "caterpillar")
plot_random_effects(fit_re, type = "qq")

# Time-varying covariates
plot_tvc_effects(fit_tvc, covariate = "stress")

# Cross-validation
plot(cv_results)
```

---

# Part X: Data Management

## Data Validation

### Pre-Analysis Checks

```{r}
validation <- validate_survey_data(
  data = mcs_simulated,
  id = "id",
  time = "age",
  outcome = "selfcontrol",
  strata = "stratum",
  cluster = "psu",
  weights = "weight",
  min_observations = 3,     # Minimum time points per person
  check_weights = TRUE,
  verbose = TRUE
)

print(validation)
```

**Checks:**
- Data structure (duplicates, missing values)
- Survey design validity (nested clusters, weights)
- Missing data patterns
- Outcome distribution
- Covariate multicollinearity

**Actions:**
- Review `validation@errors` for critical issues
- Consider `validation@warnings` for potential problems
- Follow `validation@recommendations`

## Data Preparation

### Wide to Long Format

```{r}
# Wide format data
wide_data <- data.frame(
  id = 1:100,
  y_t1 = rnorm(100),
  y_t2 = rnorm(100),
  y_t3 = rnorm(100)
)

# Convert to long
long_data <- wide_to_long(
  data = wide_data,
  id_var = "id",
  outcome_vars = c("y_t1", "y_t2", "y_t3"),
  time_values = c(0, 1, 2)
)
```

### Missing Data

surveymixr uses **Full Information Maximum Likelihood (FIML)** by default:
- Handles missing data automatically
- No need for imputation
- Valid under MAR (Missing At Random) assumption

**Patterns:**
```{r}
# Check missing patterns in validation
validation@info$missing_patterns
```

---

# Part XI: Software Integration

## Integration with Mplus

### Convert Mplus Syntax to R

```{r}
# Read Mplus .inp file
r_code <- mplus_to_surveymixr("my_analysis.inp")

# View suggested R code
cat(r_code)
```

### Generate Mplus-Style Output

```{r}
# Create Mplus-formatted output
surveymixr_to_mplus(fit, file = "surveymixr_results.txt")
```

### Validate Against Mplus

```{r}
# Compare results (manual)
compare_with_mplus(fit, "mplus_output.out")
```

**Typical Differences:**
- Minor numerical differences due to optimization algorithms
- surveymixr may find better solutions with more random starts
- Standard errors should be very similar

---

# Part XII: Simulation & Testing

## Simulate Data

### Basic Simulation

```{r}
sim_data <- simulate_gmm_survey(
  n_individuals = 1000,
  n_times = 5,
  n_classes = 3,
  time_scores = c(0, 1, 2, 3, 4),

  # Growth parameters by class
  class_params = list(
    class1 = list(intercept = 50, slope = 2),
    class2 = list(intercept = 40, slope = 0),
    class3 = list(intercept = 60, slope = -1)
  ),

  # Class proportions
  class_proportions = c(0.4, 0.3, 0.3),

  # Survey design
  design = "stratified_cluster",
  n_strata = 4,
  n_clusters = 50,

  # Missing data
  missing_rate = 0.15,

  seed = 123
)
```

### Power Analysis

```{r}
# Simulate multiple datasets to assess power
results <- replicate(100, {
  sim <- simulate_gmm_survey(...)
  fit <- gmm_survey(sim, ...)
  classification_quality(fit)$entropy
}, simplify = FALSE)

# Analyze power
mean(sapply(results, function(x) x > 0.80))
```

---

# Part XIII: Best Practices

## Workflow Recommendations

### 1. Exploratory Phase

```{r}
# Quick exploration with fewer starts
fit_explore <- gmm_survey(
  ...,
  starts = 100,
  cores = 4
)
```

### 2. Model Selection Phase

```{r}
# Comprehensive comparison
selection <- gmm_select(
  ...,
  classes = 1:6,
  criteria = c("BIC", "BLRT", "entropy"),
  blrt_samples = 100,
  starts = 200,
  cores = 4
)
```

### 3. Final Model Phase

```{r}
# Best model with more starts
fit_final <- gmm_survey(
  ...,
  n_classes = best_k,
  starts = 1000,
  cores = 4,
  keep_data = TRUE
)

# Thorough diagnostics
diagnose_convergence(fit_final)
diagnose_influence(fit_final)
classification_quality(fit_final)
```

## Reporting Results

### Essential Information

Include in publications:
1. **Model specification**: Classes, growth model, covariates
2. **Sample**: N, time points, missing data handling
3. **Survey design**: Strata, clusters, weights
4. **Estimation**: Random starts, convergence
5. **Model selection**: Criteria used, comparison table
6. **Classification quality**: Entropy, AvePP
7. **Parameters**: Growth parameters with SEs
8. **Diagnostics**: Convergence checks

### Example Results Table

```{r}
# Create comparison table
comparison <- data.frame(
  Classes = 1:5,
  LogLik = sapply(model_list, logLik),
  AIC = sapply(model_list, AIC),
  BIC = sapply(model_list, BIC),
  Entropy = sapply(model_list, function(m) entropy(m))
)

print(comparison)
```

## Computational Efficiency

### Parallel Processing

```{r}
# Use all available cores
fit <- gmm_survey(
  ...,
  cores = parallel::detectCores() - 1  # Leave 1 for system
)
```

### Memory Management

```{r}
# Don't store data unless needed
fit <- gmm_survey(..., keep_data = FALSE)

# Clear old models
rm(fit_old)
gc()
```

---

# Part XIV: Troubleshooting

## Common Issues

### 1. Convergence Failures

**Problem:** "Model did not converge"

**Solutions:**
- Increase `max_iter` (e.g., 1000 → 2000)
- Increase `starts` (e.g., 200 → 500)
- Try simpler model (fewer classes)
- Check for data issues (outliers, missing)

### 2. Separation Warnings

**Problem:** "Classes are perfectly separated"

**Solutions:**
- Reduce number of classes
- Check for extreme values or data errors
- Consider different growth model
- Add constraints on parameters

### 3. Local Maxima

**Problem:** Best solution replicated infrequently

**Solutions:**
- Increase random starts
- Use different initialization
- Try multiple runs and compare

### 4. Negative Variance Estimates

**Problem:** "Negative variance detected"

**Solutions:**
- Fix variance to positive value
- Constrain variances across classes
- Check model identification

### 5. Slow Performance

**Problem:** Estimation takes too long

**Solutions:**
- Reduce `starts` for initial exploration
- Use more `cores` for parallelization
- Start with simpler model
- Consider smaller subset for testing

## Error Messages

### Data-Related Errors

```
Error: ID variable not found
```
→ Check spelling of `id` parameter

```
Error: Duplicate id-time combinations
```
→ Data has repeated observations; fix data structure

### Estimation Errors

```
Error: Numerical instability
```
→ Try rescaling outcome variable (z-score)

```
Warning: Hessian is not positive definite
```
→ Model may be misspecified or unidentified

---

# Part XV: Advanced Topics

## Custom Growth Models

### Piecewise Linear Growth

```{r}
# Create piecewise time variable
data$time1 <- ifelse(data$time <= 5, data$time, 5)
data$time2 <- ifelse(data$time > 5, data$time - 5, 0)

# Fit model with both slopes
# (requires manual specification)
```

### Cohort Effects

```{r}
# Include cohort as time-invariant predictor in R3STEP
r3step_results <- r3step(
  gmm_object = fit,
  distal_vars = c("cohort", "other_outcomes"),
  data = distal_data
)
```

## Multiple Group Analysis

### By Gender

```{r}
# Fit separate models
fit_male <- gmm_survey(data_male, ...)
fit_female <- gmm_survey(data_female, ...)

# Compare class structures
compare_classes(fit_male, fit_female)
```

## Latent Transition Analysis

*Note: Not yet implemented. Planned for future version.*

## Growth Mixture SEM

*Note: For complex SEM with multiple processes, consider integration with lavaan.*

---

# Part XVI: References & Resources

## Key References

### Growth Mixture Modeling

1. Muthén, B. (2004). Latent variable analysis: Growth mixture modeling and related techniques for longitudinal data. In D. Kaplan (Ed.), *Handbook of quantitative methodology for the social sciences* (pp. 345-368). Sage.

2. Nagin, D. S. (2005). *Group-based modeling of development*. Harvard University Press.

### Survey Methodology

3. Lumley, T. (2010). *Complex surveys: A guide to analysis using R*. Wiley.

4. Pfeffermann, D., & Sverchkov, M. (1999). Parametric and semi-parametric estimation of regression models fitted to survey data. *Sankhyā*, 61(1), 166-186.

### Model Selection

5. Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the number of classes in latent class analysis and growth mixture modeling: A Monte Carlo simulation study. *Structural Equation Modeling*, 14(4), 535-569.

### R3STEP Method

6. Asparouhov, T., & Muthén, B. (2014). Auxiliary variables in mixture modeling: Three-step approaches using Mplus. *Structural Equation Modeling*, 21(3), 329-341.

7. Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved three-step approaches. *Political Analysis*, 18(4), 450-469.

## Getting Help

### Documentation

```{r}
# Package help
help(package = "surveymixr")

# Function help
?gmm_survey
?gmm_select

# Vignettes
browseVignettes("surveymixr")
```

### Online Resources

- **Website**: https://siyangni.github.io/surveymixr/
- **GitHub**: https://github.com/siyangni/surveymixr
- **Issues**: https://github.com/siyangni/surveymixr/issues

### Citing surveymixr

```{r}
citation("surveymixr")
```

---

# Appendix A: Function Quick Reference

## Core Functions

- `gmm_survey()` - Main estimation function
- `gmm_survey_binary()` - Binary outcomes
- `gmm_survey_ordinal()` - Ordinal outcomes
- `gmm_survey_count()` - Count outcomes
- `gmm_survey_re()` - Random effects
- `gmm_survey_tvc()` - Time-varying covariates

## Model Selection

- `gmm_select()` - Comprehensive model comparison
- `gmm_cv()` - Cross-validation
- `lmr_test()` - Lo-Mendell-Rubin test
- `lmr_sequential()` - Sequential LMR testing

## Diagnostics

- `diagnose_convergence()` - Convergence checks
- `diagnose_influence()` - Influential cases
- `diagnose_separation()` - Separation detection
- `residual_diagnostics()` - Residual analysis
- `classification_quality()` - Classification metrics
- `entropy()` - Classification entropy

## Visualization

- `plot_trajectories()` - Static trajectory plots
- `plot_interactive()` - Interactive plots
- `plot_model_selection()` - Model comparison plots
- `plot_random_effects()` - Random effects plots
- `plot_tvc_effects()` - TVC effect plots

## Utilities

- `validate_survey_data()` - Data validation
- `simulate_gmm_survey()` - Data simulation
- `wide_to_long()` - Format conversion
- `mplus_to_surveymixr()` - Mplus conversion

# Appendix B: Formulas & Mathematics

## EM Algorithm

surveymixr uses the Expectation-Maximization (EM) algorithm:

**E-Step:** Compute posterior probabilities
$$P(C_i = k | Y_i) = \frac{\pi_k f(Y_i | \theta_k)}{\sum_{j=1}^K \pi_j f(Y_i | \theta_j)}$$

**M-Step:** Update parameters
$$\theta^{(t+1)} = \arg\max_\theta \sum_{i=1}^N \sum_{k=1}^K P(C_i=k|Y_i;\theta^{(t)}) \log f(Y_i|\theta_k)$$

## Survey-Adjusted Standard Errors

Sandwich estimator:
$$\text{Var}(\hat{\theta}) = H^{-1} \Sigma H^{-1}$$

Where:
- $H$ = Hessian matrix
- $\Sigma$ = Survey-adjusted score covariance

## Information Criteria

$$\text{AIC} = -2\log L + 2p$$
$$\text{BIC} = -2\log L + p\log N$$
$$\text{aBIC} = -2\log L + p\log\left(\frac{N+2}{24}\right)$$

---

**End of User Manual**

*For updates and additional resources, visit: https://siyangni.github.io/surveymixr/*
